Heterogeneous responses to embryonic critical period perturbations within the Drosophila larval locomotor circuit

Version 1

Abstract

As developing neural circuits become functional, they undergo a phase of heightened plasticity that facilitates network tuning in response to intrinsic and/or extrinsic stimuli. These developmental windows are termed critical periods (CPs), because perturbations during, but not outside the CP, can lead to lasting and significant changes, such as the formation of sub-optimal or unstable networks. How separate, but connected elements, within a network might respond differently to a CP perturbation is not well understood. To study this, we used the locomotor network of the Drosophila larva as an experimental model, using heat stress as an ecologically relevant CP stimulus. We show that increasing ambient temperature elevates locomotor network activity. When applied during the embryonic CP, heat stress leads to the formation of a network that has suboptimal output; causing larvae to crawl more slowly and requiring longer to recover from electroshock-induced seizures, indicative of decreased network stability. Within the central nervous system, we find transient embryonic CP perturbation leads to increased synaptic drive from premotor interneurons to motoneurons, which in turn adopt reduced excitability. In contrast, the peripheral neuromuscular junction, maintains normal synaptic transmission, despite significant structural changes of synaptic terminal overgrowth and altered postsynaptic receptor field composition. Overall, our data demonstrate that connected elements within a network respond differentially to a CP perturbation. Our results suggest a sequence, or hierarchy, of network adjustment during developmental CPs, and present the larval locomotor network as a highly tractable experimental model system with which to study CP biology.

Introduction

During nervous system development, critical periods (CPs) are transient windows of heightened plasticity, which are fundamental to the emergence of appropriate network function. As a consequence, networks are particularly vulnerable to activity perturbations during a CP, which can result in lasting structural and/or functional changes, while comparable perturbations outside the CP are less impactful (Coulson et al., 2022; Reh et al., 2020). The functions and concepts of CPs have traditionally been investigated in the visual system of mammals (Hensch, 2005; Hensch and Quinlan, 2018; Hubel and Wiesel, 1970) and imprinting of birds (Hess, 1975; Horn, 2004). More recently, CPs have also been identified and investigated across a range of species and neuronal ensembles, from networks underlying the development of speech (Harrison et al., 2005) to motor behaviour of rats (Soiza-Reilly and Azcurra, 2009), fish (Hageter et al., 2023) and insects (Dombrovski and Condron, 2021; Giachello and Baines, 2015). The presence of a CP in developing networks is therefore increasingly considered a universal phenomenon. The precise changes within networks and associated mechanisms, as they occur during normal development, as well as those resulting from perturbations during a CP, are not well understood. This is in part due to the complexity of many of the experimental model systems, notably mammalian cortical systems that have been the predominant model for studying CPs.

To overcome this technical obstacle, we have utilized the comparatively simpler nervous system of the fruit fly larva, Drosophila melanogaster, with which to study CPs. Though an insect, Drosophila displays CPs with features that suggest commonalities to mammalian CPs (for review see: Coulson et al., 2022). Currently, the best characterised CP during Drosophila development is one that occurs during embryogenesis, linked to the emergence of normal locomotor network output (Carreira-Rosario et al., 2021; Corke et al., 2025; Coulson et al., 2022; Crisp et al., 2008, 2011; Ackerman et al., 2021; Fushiki et al., 2013; Giachello et al., 2021; Giachello and Baines, 2015; Hunter et al., 2024; Williamson et al., 2021; Zeng et al., 2021). This experimental system benefits from the comprehensively annotated connectome of the Drosophila larval nervous system (Giachello et al., 2020; Heckman and Doe, 2022; Valdes-Aleman et al., 2021; Winding et al., 2023) as well as an unparalleled set of genetic tools for manipulating identified neurons with both spatial and temporal precision (Simpson, 2009). Combined with imaging and electrophysiological techniques, this model system enables investigation of cellular responses to CP manipulations at the level of identified cells and synapses (Carreira-Rosario et al., 2021; Coulson et al., 2022; Crisp et al., 2008, 2011; Fushiki et al., 2013; Giachello et al., 2021; Giachello and Baines, 2015; Hunter et al., 2024; Williamson et al., 2021; Zeng et al., 2021).

Neuronal activity manipulations during the CP of Drosophila embryos have been shown to cause decreased network stability. This manifests in significantly longer times required for larvae to recover from a strong activity challenge, such as electric shock-induced seizures (Coulson et al., 2022; Giachello et al., 2021; Giachello and Baines, 2015; Hunter et al., 2024, 2021). The premise that balanced network activity during the CP is necessary for stable, resilient networks to form has been explicitly supported by findings from our recent studies (Giachello et al., 2021; Giachello and Baines, 2015; Hunter et al., 2024). By extension, rectifying activity manipulations during identified CPs could have therapeutic potential (Nardou et al., 2023).

While experimental perturbations of neuronal activity during development have been a useful methodological approach for studying CPs as a phenomenon and the associated biological mechanisms, such manipulations might seem artificial in the context of normal development. Most animals are not warm-blooded and in poikilothermic animals, such as Drosophila, changes in ambient temperature, daily as well as seasonal, lead to changes in behavioural activity, likely caused by changes in neuronal and network activity (Lee and Montell, 2013; Powsner, 1935). Studies from the crustacean stomatogastric nervous system showed that these minimal networks produce a characteristic, rhythmic firing pattern, which is robustly maintained across a range of external temperatures. Importantly, neuronal network properties were adapted relative to the long-term temperatures that crabs had experienced, resulting in significantly different upper “crash temperatures” at which networks ceased to function (Marder, 2023; Marder and Rue, 2021). In Drosophila, the temperature experienced by late embryos determines larval locomotor output, including crawling speed and social feeding behaviour (Williamson et al., 2021). Similarly, studies on the Drosophila adult visual and olfactory systems showed that neuronal wiring properties and associated behaviours of adults are determined by the temperature experienced during a second CP of Drosophila, which takes place during pupariation (Chodankar et al., 2020; Kiral et al., 2021; Züfle et al., 2025), a phase of extensive network remodelling (Baumann et al., 2024; Leier et al., 2025; Lowe et al., 2023; Nelson et al., 2024).

In this study, we investigated ambient temperature as an ecologically relevant stimulus. To gain insights at the level of single, identified connected cells, we focused on the neuromuscular system of the Drosophila larva, and discovered that the components of this network respond differently to transient heat stress, when experienced during the CP. Specifically, we find that elevated temperature increases neuronal activity, and that transient heat stress during the locomotor CP causes long term reduced network stability, similar to the effects of neuronal activity manipulations (Giachello et al., 2021; Giachello and Baines, 2015; Hunter et al., 2024). These changes are accompanied by presynaptic terminal overgrowth at the neuromuscular junction (NMJ) and changes in postsynaptic neurotransmitter receptor composition, though synaptic transmission is homeostatically regulated and remains normal. In contrast, within central circuitry, we find that transient heat stress leads to reduced excitability of motoneurons, potentially as a response to enhanced synaptic drive from premotor interneurons. Overall, this study presents a first comprehensive analysis of how several interconnected elements of a defined locomotor network develop in response to CP perturbation by heat stress, an ecologically relevant stimulus.

Results

Temperature manipulation during the locomotor CP leads to an unstable crawling network

We had previously identified a CP of nervous system development in late embryogenesis, from 17 – 19 hrs after egg laying (AEL), equivalent to 80-90% of embryonic development (Giachello and Baines, 2015). Perturbation of network activity during this developmental time window, pharmacologically or optogenetically, leads to the formation of an unstable, seizure prone network. These phenotypes manifest most clearly upon acute activity challenges. For example, an electric shock administered to late larvae, that during their embryonic CP had experienced an activity manipulation, causes seizures with recovery times that are significantly longer compared to unmanipulated controls. Because large-scale activity manipulations during the embryonic CP are artificial, We asked to what extent changes in environmental conditions, that embryos might normally encounter, might also alter network development. As poikilothermic animals, Drosophila development strongly depends on ambient temperature (Powsner, 1935). Several studies have shown that temperature experiences during development alter neuronal wiring as well as behaviours in larval and adult flies (Kiral et al., 2021; Williamson et al., 2021; Züfle et al., 2025).

We tested heat stress of 32°C (Scialò et al., 2015), which is known to be experienced by Drosophila melanogaster (Williamson et al., 2021), i.e. 7°C above the evolved temperature preference of 25°C (Ashburner et al., 1983). We exposed embryos to 32°C heat stress before, during, or after the embryonic CP, then returned the embryos to the preferred temperature of 25°C. After four days, at the end of larval development, we measured the speed of recovery from electric shock-induced seizures, as a proxy for network stability. Transient heat stress experienced before or after the embryonic CP did not result in a change from unmanipulated controls (i.e. no change in seizure response), while in animals that had experienced 32°C heat stress during the embryonic CP (i.e. between 17 – 19 hrs) had significantly prolonged recovery times from induced seizure, indicative of reduced network stability (Fig. 1A). Several studies had previously shown that increasing temperature leads to increased activity in the locomotor network (Oswald et al., 2018; Sigrist et al., 2003; Zhong and Wu, 2004). We tested this directly by functional imaging in isolated CNS, measuring spontaneous calcium transients from motoneurons. Specifically, we expressed GCaMP8f selectively in aCC motoneurons using R94G06-GAL4. Stepwise increases in temperature led to both, an increase in the GCaMP signal amplitude and in the frequency of rhythmic bursts. In contrast, CNSs kept at the control temperature of 25°C for the duration of functional imaging tended to show a decreasing GCaMP signal amplitude (Fig. 1B). These experiments suggest that increases in ambient temperature, above 25°C, lead to concomitant increases in locomotor network activity. When those that occur during the embryonic CP, result in altered developmental outcomes, notably of sub-optimal, less stable networks.

Transient heat stress during the critical period leads to unstable networks.

A Schematic representation of embryonic and larval development and temperature manipulation during embryogenesis. Embryos were kept at 32°C for 2 h either before (13-15 h, 15-17 h AEL), during (17-19 h AEL) or after (19-21 h AEL) the locomotor network CP. B Acute increase of temperature leads to increased network activity, as measured with the fluorescent calcium indicator GCaMP8f expressed in segmentally repeated aCC motoneurons in isolated central nervous systems; while constant 25°C leads to reduced activity over time.

Temperature manipulation during the embryonic CP leads to changes in larval NMJ morphology and composition

To investigate structural and functional changes that result from a CP heat stress perturbation we first focused on the well characterised, and easy to access, peripheral neuromuscular junction (NMJ). Specifically, we focused on the NMJ of dorsal acute muscle 1 (DA1), which is innervated by the aCC motoneuron, a cell that has been the focus of our previous studies centred around CP perturbations (Fig. 2A) (Baines et al., 2001, 1999; Fujioka et al., 2003; Hunter et al., 2024; Landgraf et al., 2003; Giachello et al., 2021; Giachello and Baines, 2015). Because temperature manipulations outside the CP did not lead to an unstable network (Fig. 1A), we adopted a simplified experimental paradigm of putting embryos from a 6h egg collection to 32°C for 24h. This produces identical outcomes to heat stress limited to the CP, and it ensures that all embryos go through their CP during heat stress irrespective of minor variations in developmental timing or due to retention of fertilised eggs. We find that embryonic heat stress leads to a significant overgrowth of the aCC motoneuron terminal (type Ib boutons) by the late larval stage, resulting in an increase in the number of boutons and active zones, while overall active zone density is not altered (Fig. 2B). On the postsynaptic side, embryonic CP heat stress causes an altered receptor composition: a selective reduction of the larger conductance GluRIIA subunit. Across the aCC NMJ, this leads to a reduced ratio of the larger conductance GluRIIA versus the lower conductance GluRIIB subunit (Fig. 2C-Cii).

Morphological changes at the neuromuscular junction after critical period perturbation.

A) Schematic of temperature manipulation during embryogenesis and location of dorsal acute muscle 1 (DA1) within a fillet dissected larva. Embryos collected over a 6h period were incubated at 32°C for 24h to guarantee heat stress exposure during the embryonic CP, 17-19 hours after egg laying. B) This results in a change of subsequent NMJ development with significant overgrowth of aCC presynaptic terminals (green) and an increase in bouton number and NMJ area. A concomitant increase in active zone number (magenta) results in a normal active zone density. C) Postsynaptically, CP heat stress result in a decrease of GluRIIA (Ci) while GluRIIB levels remain unaffected (Cii).

Despite morphological changes, larval NMJ physiology is not altered

In view of the prominent changes in NMJ growth, morphology and postsynaptic glutamate receptor composition that result from heat stress (32°C) during the embryonic CP, we next assessed synaptic physiology at this NMJ. We used the two-electrode voltage clamp (TEVC) technique, initially measuring single pulse transmission across a gradient of external calcium concentrations. Despite the marked structural changes in manipulated animals, single pulse transmission was not significantly different between embryonically manipulated (32°C experience during their CP) and controls that experienced 25°C throughout (Fig. 3 A&B). Quantification of the number of release sites utilised per action potential and quantal size (Jusyte et al., 2023) also showed no statistically significant difference between CP-manipulated and control specimens (Fig. 3C). Next, we quantified the size of the readily releasable pool of synaptic vesicles by recording responses to a high frequency train of action potentials (Fig. 3D), then plotting the cumulative amplitude of responses to the first 60 pulses and linearly back extrapolating through the last 20 pulses to y=0 (Fig. 3E). This also did not show a significant difference between CP-manipulated and control animals (Fig. 3F). Though we noticed that during the 60Hz pulse train, NMJs in CP-manipulated specimens exhibited an increased facilitation during the first 10 pulses, indicating a slight reduction in release probability of the presynaptic terminal (Fig. 3D). Last, we tested if the NMJ remained able to adjust homeostatically to acute changes in synaptic transmission. We utilized the well-established method of acutely blocking glutamate receptors with Philanthotoxin (PhTx), a GluRIIA blocker, to induce presynaptic homeostatic potentiation (Frank et al., 2006; Genç and Davis, 2019; Müller and Davis, 2012). Acute GluRIIA blockade by a 10 min incubation with PhTx caused a significant reduction mEPSC amplitude while evoked EPSC amplitude remained unchanged in both controls and CP-manipulated specimens. Calculating quantal content shows that in response to acute GluRIIA blockade, both control and CP-manipulated animals are indeed able to induce presynaptic homeostatic potentiation by increasing quantal content (Fig. 3G-Giii). While establishing baseline conditions, we found mEPSC amplitudes and quantal content to be comparable between controls and CP-manipulated animals Fig. 3G-Giii). This was contrary to expectation in light of the overall reduction in the GluRIIA/B ratio that results from a CP heat stress manipulation. However, we noticed that reductions in GluRIIA signal were most pronounced in proximal regions of the NMJ; while distal boutons, which have a higher probability of release under low frequency stimulation regimes (Peled and Isacoff, 2011), have normal levels of GluRIIA. The preferred recruitment of distal boutons for release could explain the apparent lack of effect on mEPSC amplitude. Overall, we conclude that despite the significant structural changes that emerge by late larval stage following a transient embryonic 32°C heat stress experience, synaptic physiology at the late larval NMJ remains normal. Moreover, when considering the unstable network phenotype, as manifest by increased recovery times from electric shock-induced seizures, our finding that synaptic transmission at the NMJ remained normal suggests that network instability likely originates from changes within the central locomotor circuitry.

Physiological NMJ properties after 32°C perturbation.

A) Representative traces of control two electrode voltage clamp recordings across a gradient of external calcium concentrations. B) Postsynaptic current amplitude is not altered between control and CP-manipulated animals at any given external calcium concentration. C) Plotting postsynaptic current amplitude against its variety to calculate quantal size (q) and the number of release sites being used (N), also did not show a significant difference between control and CP-manipulated animals. To measure readily releasable pool size, a 60Hz stimulus train was applied for 1s (D) where we could observe a slight increase in facilitation over the first 10 pulses in specimens that in their embryonic stage had experienced heat stress of 32°C. Cumulative amplitude was plotted over time and a linear regression was back extrapolated through the last 20 pulses (E). The y intercept was divided by the Mini size of the same trace to calculate the readily releasable pool size. G-Giii) NMJs in CP-manipulated specimens have normal EPSC and mEPSC amplitudes and quantal content. Acute application of the GluRIIA blocker PhTx reliably induced presynaptic homeostatic potentiation in controls and 32°C CP-manipulated animals.

Motoneurons display reduced excitability and receive increased synaptic input following CP perturbation by heat stress

We noted that in CP-manipulated larvae, motor axons required increased stimulation voltage to reliably induce action potentials. This hints at motoneurons in CP-manipulated larvae having acquired reduced excitability. To directly test motoneuron excitability, we used whole cell patch clamp to measure evoked action potential firing, as a measure of excitability of the aCC motoneuron. These measures were made in the presence of mecamylamine (200 µM in external saline) to block endogenous cholinergic excitatory synaptic drive. Action potential firing of aCC was evoked using injection of current steps (4pA). We found that in late L3, aCC had significantly decreased excitability following an embryonic CP experience of 32°C heat stress (Fig. 4A). These results are comparable to, and validate, results obtained after optogenetic locomotor network activity manipulation during the same embryonic CP (Giachello and Baines, 2015). In previous studies, we had shown with recordings from aCC that its synaptic drive is enhanced following transient CP manipulations of network activity, using optogenetic or pharmacological means to induce network over-excitation, or genetic overactivation of neurons, as in parabss seizure mutants (Giachello and Baines, 2015; Hunter et al., 2024). Specifically, endogenous excitatory cholinergic synaptic currents, termed spontaneous rhythmic currents (SRCs), were increased in duration, though not in amplitude, following network activity manipulations during the CP (Giachello et al., 2019; Giachello and Baines, 2015). We asked if heat stress during the CP led to similar changes. Indeed, in late larvae that had transiently experienced heat stress during their embryonic CP, SRCs recorded from aCC motoneurons were significantly increased in duration, though with amplitude unchanged (Fig. 4Bi). Note these SRC recordings were undertaken in the absence of mecamylamine and suggest an increase in premotor network activity. Together with the decrease in membrane excitability of aCC, these observations are compatible with the idea that motoneurons might have undergone homeostatic adjustment in response to an increase in premotor excitatory synaptic drive caused by the CP manipulation.

Larval motoneurons display reduced excitability and increased synaptic input after an embryonic 32°C CP-manipulation.

A aCC motoneurons at the late third instar larval stage, isolated from excitatory synaptic input. Motoneurons in specimens, which during the embryonic CP had transiently experienced 32°C heat stress, show significant reduction in excitability relative to controls. B Left: whole cell patch clamp recordings from the aCC motoneuron reveal an increase in spontaneous rhythmic current duration several days after embryos had experienced transient 32°C heat stress during the CP. Right: targeted optogenetic activation (using Chronos) of a premotor excitatory interneuron (A27h) showed that A27h input is increased in late larvae that as embryos had experienced 32°C heat stress during the CP.

To test that hypothesis, we recorded synaptic drive to aCC, under whole cell patch clamp configuration, while optogenetically activating a specific and strongly connected excitatory premotor interneuron, termed “A27h” (Hunter et al., 2023). Because motoneurons are quiescent without synaptic drive, a small amount of current was injected to aCC to maintain a modest and consistent mean firing frequency of ∼10-20 Hz. As might be expected, and as per our previous study (Hunter et al., 2024), acute optogenetic activation of the A27h premotor excitor resulted in increased aCC firing frequency. Significantly, this increase was greater in larvae that had been exposed to heat stress during their embryonic CP, compared to controls (Fig 4.Bii). This observation supports the hypothesis that heat stress during the embryonic CP results in increased excitatory drive from central interneurons, which is then countered by postsynaptic motoneurons through reduction in their excitability.

Homeostatic adjustments of the aCC motoneuron result in normal endogenous firing patterns

If the adjustments displayed by the aCC motoneuron are indeed a homeostatic response, then one might predict this to be a mechanism to maintain the endogenous level of motoneuron activity. To test this, we used whole cell patch recordings from aCC motoneurons, in isolated nerve cords of late third instar larvae. As hypothesized, our recordings show no significant differences in endogenous motoneuron activity as a result of embryonic CP heat stress, measured by burst rate, burst duration, or action potentials per burst (Fig. 5A). Finally, to test if homeostatic adjustments of motoneurons translate to normal muscle output, we recorded spontaneous synaptic drive from muscles in a semi-intact preparation with an intact nervous system attached. Current clamp recordings from the DA1 muscle, while supressing muscle contractions using nifedipine (50µM) (Dyson et al., 2022; Kratschmer et al., 2021), showed no significant differences in spontaneous burst frequency or duration, nor the overall time spent bursting (Fig. 5B), between controls and CP heat stress-manipulated specimens. Therefore, we conclude that CP manipulations act primarily on the cells of the premotor circuitry, namely excitatory premotor interneurons as tested in this study. Changes in the motoneurons can be seen to constitute secondary homeostatic adjustments in response to those premotor network changes.

Spontaneous larval motoneuron and muscle firing properties and are unchanged after embryonic temperature challenge.

A Monitoring spontaneous firing by loose cell patch clamp recordings from the aCC motoneuron in a late third instar larval nerve cord did not show a significant difference between CP-manipulated specimens and controls. B Measuring muscle output of motoneuron spontaneous firing by current clamp recordings from the DA1 muscle in late third instar larvae showed that animals are able to produce normal motoneuron firing output to the muscle, irrespective of embryonic experience.

Inter-segment wave propagation speed is reduced following embryonic critical period perturbation

To test how excitatory premotor interneurons are affected following a transient embryonic 32°C CP manipulation, we expressed the calcium indicator GCaMP8f in A27h interneurons, and in isolated 3rd instar nerve cords recorded the propagation of neuronal activation waves across abdominal segments, indicative of fictive locomotion (Fig.6A) (Streit et al., 2016; Pulver et al., 2015). These experiments show that a transient 32°C exposure during the embryonic CP leads to a significant increase in the time required for this neuronal activity wave propagation, as compared to controls. This mirrors a decrease in crawling speed in intact larvae (Fig.6B). Because our electrophysiological analyses showed that motoneuron firing patterns and output to the muscle remain normal after a 32°C CP manipulation (Fig.3-5), these results further support the hypothesis that it is the lasting changes to the central premotor circuitry, rather than to the motoneurons, which are responsible for the changes in crawling behaviour, and likely also for network instability.

CP manipulation causes reduced speed of activity wave propagation and larval crawling.

A Functional imaging of A27h premotor interneurons using GCaMP8f in isolated CNSs from late larvae that previously had experienced transient 32°C heat stress during their embryonic CP. Propagation of the activity wave is slowed down, due to an increase in intersegment duration, notably between anterior segments, e.g. A3 and A1. B In intact animals, network activity is similarly altered: CP heat stress (32°C) leads to a significant reduction in larval crawling speed, as compared to controls (25°C).

Discussion

An experimental model system with which to study CP biology

In this study we have utilised the larval neuromuscular system as an experimental model with which to study the role of CP-regulated plasticity across interconnected cells within a defined circuit. We have identified 32°C heat stress as an ecologically relevant CP stimulus: transient exposure of embryos to 32°C during the previously defined CP for the developing CNS (17-19 hours after egg laying) leads to lasting changes in subsequent nervous system development, resulting in reduced network stability which is reflected by an increased susceptibility to electro-shock induced seizures. As is characteristic for CPs, the same manipulation before or after this defined developmental window has no lasting effect on nervous system properties or animal behaviour. Calcium imaging of motoneurons showed that stepwise increases in ambient temperature cause acute increases of rhythmic locomotor network activity. This is in line with past studies, which had demonstrated the well-established link between temperature and neuronal activity in this same experimental system (Zhong and Wu, 2004). It is perhaps, therefore, not surprising that 32°C heat stress during the embryonic CP of nervous system development causes outcomes that are comparable to neuronal activity manipulations (pharmacological or genetic) during this developmental window (Giachello et al., 2021; Giachello and Baines, 2015; Hunter et al., 2024).

Different cell types respond differently to CP manipulations

A particular strength of the Drosophila larval motor system is the accessibility of interconnected network elements. We focused on a small part of the locomotor network: an excitatory premotor interneuron (A27h) connected to a motoneuron (aCC), and its postsynaptic target body wall muscle (DA1). Following transient application of heat stress as an ecologically relevant during the embryonic CP, we could identify that within this section of the locomotor network each of these three connected nodes responded differently. We observed increased excitatory drive from A27h interneurons to aCC motoneurons, while those motoneurons displayed homeostatic adaptation; namely, a decrease in excitability, leading to maintenance of normal firing patterns. At the presynaptic NMJ terminal in the periphery, in contrast, we documented normal synaptic transmission properties, despite a significant presynaptic terminal overgrowth. The latter might be a secondary adjustment to a decrease in transmitter sensitivity by the muscle, which showed reduced GluRIIA levels along parts of the NMJ terminal.

The first observation to make is that the responses have been reproducible. This suggests that there might be direct cause-effect relationships at the cellular level, which can now be investigated in detail. An alternative interpretation is that activity patterns in this part of the network might be robustly canalised to affect reproducible outcomes. Second, the changes in neuronal excitability and synaptic transmission are compatible with a temporal sequence, or a functional hierarchy, during network maturation. For example, following an embryonic heat stress, premotor interneurons, potentially part of the central pattern generator (Carreira-Rosario et al., 2021; Zeng et al., 2021), increase their collective synaptic drive to postsynaptic motoneurons, which in turn reduce their excitability (Figure 4). Thus, the changes in motoneuron properties that we have characterised, we interpret as homeostatic responses to the increase in presynaptic excitatory drive from premotor interneurons. This implies a hierarchy, whereby the specification of premotor interneuron properties dominates that of their downstream integrators, the motoneurons. A sequence of cell type-specific maturation windows has recently been shown in the developing mouse neocortex. There, the somatostatin-expressing inhibitory interneurons differentiate early and promote spontaneous activity through inhibition of the later differentiating parvalbumin-expressing inhibitory interneurons (Mòdol et al., 2024).

NMJ structural development is affected by CP manipulations but not synaptic transmission

Despite embryonic heat stress leading to significant NMJ structural changes, in terms of presynaptic terminal overgrowth (an average increase of active zone number by 30%) and overall reduction by 50% of GluRIIA, evoked transmission resulting from the firing of single action potentials in the motor axon remained normal. Closer inspection suggested that GluRIIA distribution was no longer homogenous in late larval NMJs if the animal had experienced heat stress during its embryonic CP. While GluRIIA levels were reduced proximally (axon entry point), distal boutons, in contrast, retained normal GluRIIA levels. Distal boutons had previously been shown to house release sites that are preferentially recruited under low frequency stimulation regimes (Peled and Isacoff, 2011). This could explain why we observed normal evoked transmission despite the significant reduction of GluRIIA at proximal sites. If this was the case, then one might expect high frequency stimulation regimes, e.g. of 60 Hz, to recruit many other presynaptic sites for vesicle fusion, including the proximal sites that are apposed to postsynaptic densities with reduced GluRIIA. Although we did not conduct functional imaging to investigate this directly, a 60 Hz stimulation regime did indeed reveal increased facilitation in animals that experienced heat stress during their embryonic CP, indicating a reduction in presynaptic release probability across the whole NMJ. Though statistically not reaching significance, changes in release probability would be compatible with the idea that at sub-60 Hz stimulation a much larger fraction of synaptic release sites are recruited, including proximal sites with reduced GluRIIA (Figure 3D). That evoked synaptic transmission at the NMJ is not significantly affected by CP manipulations, is perhaps to be expected. A well-established feature of this synapse is that its transmission is tightly regulated to maintain consistency. The various NMJ plasticity mechanisms include presynaptic homeostatic potentiation (PHP), which we have shown to function effectively even after CP heat stress manipulation (Delvendahl and Müller, 2019; Frank et al., 2020, 2006; Goel and Dickman, 2021).

It is unlikely that structural changes at the NMJ contribute to making the nervous system less stable, or that they effect CP-regulated changes to larval behaviour, such as crawling speed or social feeding (Williamson et al., 2021). Based on previous CP manipulations targeted to various neuronal subsets (Giachello et al., 2021), it seems more likely that behavioural phenotypes, including reduced crawling speed and increased recovery times from electric shock-induced seizures, would be the consequence of changes in the central circuitry. Indeed, functional imaging of the excitatory premotor interneuron, A27h, showed that embryonic heat stress leads to networks with significantly slower intersegmental propagation of fictive locomotor waves, compatible with the reduced crawling speed seen in intact larvae (Figure 6). This contrasts with acute heat stress causing an increase in such fictive locomotor waves of activity (Figure 1B). Heat stress during the CP could cause a malfunctioning of pacemaker activity. Alternatively, the observed slowing of the CPG in larval stages could be a network-wide homeostatic response to an abnormally fast rate experienced due to heat stress exposure during the embryonic CP. We propose that the NMJ, because it is a large, readily accessible peripheral synapse whose development is altered following transient CP heat stress, could be serve as an experimental model for studying CP-associated mechanisms.

Materials and Methods

Fly husbandry

Wild-type and transgenic strains were maintained on standard yeast–agar–cornmeal medium at 25°C. For CP manipulations, animals were kept in laying pots with apple juice agar plates and yeast paste for 6h. Collected eggs were either kept at 25°C for controls or at 32°C for 24h, before being moved back to 25°C until animals developed to wandering L3 larvae.

Dissection and Immunostainings

Wandering L3 larvae were dissected in Sørensen’s sodium phosphate buffered saline (0.1 M, pH 7.2) and fixed for 10min at room temperature with Bouińs fixative (Sigma-Aldrich). After fixing, specimens were washed in washing solution (Sørensen’s buffer + 0,25% BSA + 0,3% TritonX-100 from Sigma-Aldrich) and incubated with mouse nc82 (anti-Bruchpilot to reveal active zones) antibody (1:100, DSHB) overnight, visualized with goat α-mouse StarRed (1:1000, Abberior, cat.no.STRED-1001-500UG, Lot. 2042PK-1) and AlexaFluor488 conjugated goat α-HRP (1:1000, Jackson ImmunoResearch, cat.no. 123-545-021, Lot. 152767) to visualise presynaptic, neuronal structures. For stainings of the postsynaptic glutamate receptor subunits GluRIIA and GluRIIB we used monoclonal mouse-α-DGluRIIA antibody (1:100, DSHB, product ID: 8B4D2 (MH2B)) and polyclonal rabbit-α-DGluRIIB antiserum (1:1000, gift from Mihaela Serpe (Sulkowski et al., 2016)) in conjunction with goat α-mouse StarRed (1:1000) and goat α-rabbit Atto594 secondary antibodies (1:1000, Hypermol, cat.no. 2306, Lot. 0356RG22), respectively, in combinations with AlexaFluor488 conjugated goat α-HRP (1:1000, Jackson ImmunoResearch, cat.no. 123-545-021, Lot. 152767) to visualise presynaptic terminals.

Fly stocks

Image acquisition

Confocal images were acquired, using a Leica Stellaris point scanning confocal microscope, with an 40x/1.25N.A. oil immersion lens (Leica) and a 1.25 digital zoom. Images were acquired with a voxel size 0.16x 0.16x 0.3 µm. Images were processed using FIJI and Affinity Designer.

Two electrode voltage clamp

Two Electrode Voltage Clamp (TEVC) recordings were performed in HL3.1 saline (Feng et al., 2004) composition in mM: NaCl 70, KCl 2.5, MGCl2 4, CaCl2 2 (if not other mentioned), NaHCO3 10, trehalose 5, sucrose 115, HEPES 5 (pH was adjusted to 7.24-7.25 with 1M NaOH), using an Olympus BX50WI compound microscope with 60X dipping (Olympus 60x/0.9 N.A.) objective lenses. Experiments were performed on third-instar larval NMJs of the dorsal acute muscle 1 (DA1) in segments A3 and A4. The respective nerve was sucked in with a self-built suction electrode (Johnson et al., 2007) which was filled with HL3.1 saline. Subsequently, the nerve was stimulated for 0.1ms to trigger action potential firing. Briefly, voltage was stepwise increased until the first synaptic transmission could be recorded. From that voltage, 1 V was added to reliably evoke APs in control conditions, in 32°C CP manipulated animals 2-3 V had to be added to reliably recruit aCC and RP2. For PhTx experiments, samples were incubated in 4 µM PhTx final concentration (Frank et al., 2006) for 10 min. While dissecting, great care was taken to not stretch animals too much. Successful PhTx application was determined by dividing miniature EPSC amplitude of PhTx treated animals by the average Amplitude of control mEPSCs. If PhTx treated mEPSCs were at least 40% reduced compared to control, PhTx application was considered successful (Frank et al., 2006). Quantal content was determined by dividing EPSC by mEPSC for each recording. Recordings were only further considered from cells with an initial Vm below -50 mV and input resistance between 8-10 MΩ, using intracellular electrodes filled with 3M KCl with resistances of ∼20 MΩ (Electrodes for current injection and the membrane voltage). Current passing und recording electrodes were pulled from Borosilicate glass microelectrodes with filament (OD1 mm, ID 0.58mm; Harvard Apparatus LTD, GC100F-10 Part No. 30-0019), suction electrodes from Borosilicate glass microelectrodes without filament (OD1 mm, ID 0.58mm; Harvard Apparatus LTD, GC100-10 Part No. 30-0016) with a Flaming/Brown Puller Model P-97 (Sutter Instruments USA). Excitatory postsynaptic currents (EPSCs) were evoked by an Isolated Pulse Stimulator (Model 2100 A-M Systems USA), recorded at a clamped voltage of -60 mV with a sampling rate of 50 kHz and filtered with the anti-alias filter at 25 kHz using the AxoClamp 2B amplifier (Molecular Devices Axon Instruments USA). Data analysis was done with Clampfit 10.7. Traces were first filtered with a 360 Hz Gaussian low pass filter, then template searches were run over the trace for single pulses and spontaneous miniature excitatory postsynaptic currents (mEPSCs or minis). The resulting files were used to analyse mEPSC and EPSC properties. Rise and decay tau were measured with an one-term exponential product fit (Levenberg-Marquadt method with a max of 5000 iterations)

Current clamp muscle recordings

Current clamp recordings from muscle DA1 were performed in HL3.1 saline (see above), in semi intact, fillet dissected wandering L3 larvae with intact nervous system. After finishing the dissection, only animals performing crawling-like movements were considered for further experiments. To block muscle contractions, a 500 µM stock solution of nifedipine in HL3.1 and 1% DMSO was added to the bath for a final concentration of 50 µM nifedipine and 0.1% DMSO for 15 min. Recordings were only further considered from cells with an initial Vm below -60 mV and input resistance between 8-10 MΩ, using intracellular electrodes filled with 3M KCl with resistances of ∼20 MΩ. Spontaneous activity was recorded for 5min. Burst analysis was performed using the Clampfit Burst Analysis, using the Poisson Surprise model with a minimum of 5 events per burst.

Electroshock assay

A handheld stimulator consisting of two tungsten wires (0.1 mm diameter) fixed to a plastic rod was used. Wires were shaped to be ∼1-2 mm apart, attached to a DS2A Mk.II stimulator (Digitimer Ltd), which was set to provide a 6 V, 2 s current pulse. Larvae were placed on a plastic plate lid, with tissue paper used to dry residual moisture from the larvae. Once larvae had returned to normal crawling, the tungsten wires were brought into contact with the larvae, positioned on the larval cuticle, perpendicular to and over the approximate position of the CNS, with gentle pressure. Current stimulation was then delivered, which usually induced a sustained contraction and paralysis representing a seizure (videos can be viewed in Marley and Baines 2011). The time to resumption of normal behaviour (regular, whole body forward propagating peristaltic waves) was recorded and defined as the duration of recovery from seizure.

Dissection and CNS mounting for calcium imaging and electrophysiology

Larval CNSs were dissected and mounted for electrophysiological recordings and for GCaMP imaging. Third instar (L3) larvae (of either sex) were dissected in a dish in standard saline (135 mM NaCl (Fisher Scientific), 5 mM KCl (Fisher Scientific), 4 mM MgCl2·6H2O (Sigma-Aldrich), 2mM CaCl2·2H2O (Fisher Scientific), 5 mM TES (Sigma-Aldrich), 36 mM sucrose (Fisher Scientific), pH 7.15), to remove the CNS (ventral nerve cord and brain lobes). The isolated CNS was then transferred to a droplet of external saline, within which it was laid flat (dorsal side up) and glued (GLUture Topical Tissue Adhesive; World Precision Instruments USA) to a Sylgard-coated cover slip (1 to 2mm depth of cured SYLGARD Elastomer (Dow-Corning USA) on a 22 x 22mm square coverslip). The preparation was then placed on a glass slide and viewed under a microscope (Olympus BX51-WI).

Calcium imaging

Dissected CNSs were imaged (QImaging EXi-Aqua; photometrics, Arizona, US) with an acquisition frame rate of 10Hz and frame duration of 100 ms using a x20 immersion lens. GCaMP8f was excited using a 470 nm collimated LED (Thorlabs, New Jersey, US). Time-course of neuronal fluorescence was determined in WinFluor (V.4.1.5; University of Strathclyde, UK) by adding regions of interest (ROIs) to axons, close to the cell body (Streit et al, 2016). In Clampfit (10.3.1.5; Molecular Devices, California, US) fluorescence traces were smoothed using a lowpass BoxCar filter using 11 smoothing points prior to analysis.

For motoneuron calcium transients (Fig.1), recordings were conducted for 600s, whilst temperature of the explanted CNS was increased steadily from ∼23°C to ∼34°C using a digital heating unit (CO 102; Linkam Scientific, Surrey, UK). For sham recordings, temperature was maintained at a constant 25°C. Temperature was measured in Clampex (10.3.1.5; Molecular Devices, California, US) using a temperature sensor placed in the external saline surrounding the brain, via a temperature controller (TC-10; npi electronic GmbH, Tamm, Germany). The frequency of calcium fluorescence peaks was measured for each segment (A8/9 – A1) in three sequential 200 s time periods and averaged per larva. Average temperature was calculated for each recording during these same 200 s time periods.

For calcium imaging of interneurons after CP manipulation (Fig.5), 180 s recordings were taken. Time lag was determined by manually assessing the time between peak fluorescence corresponding to ipsilaterally adjacent interneurons (segments A6 – A1. For each recording the first complete forward waves of activity were used (up to three repeats per trace) to generate an average intersegment duration for each pair of interneurons.

Whole cell patch clamp electrophysiology

CNSs were dissected, mounted (see calcium imaging methods) and viewed under a 60x water-immersion lens. To access soma, 1% protease (Streptomyces griseus, Type XIV, Sigma-Aldrich, in external saline) contained within a wide-bore glass pipette (GC100TF-10; Harvard Apparatus UK, approx. 10 µm opening) was applied to abdominal segments, roughly between A5-A2 (Baines and Bate, 1998). This was done to remove overlaying glia to facilitate access to underlying nerve cell soma (e.g., aCC). Motoneurons were identified by anatomical position and relative cell size, with aCC being positioned close to the midline and containing both an ipsilateral and contralateral projection. Recordings were made using borosilicate glass pipettes (GC100F-10, Harvard Apparatus) that were fire polished to resistances of 10 - 15MΩ when filled with intracellular saline (140 mM potassium-D-gluconate (Sigma-Aldrich), 2 mM MgCl2·6H2O (Sigma-Aldrich), 2 mM EGTA (Sigma-Aldrich), 5 mM KCl (Fisher Scientific), and 20 mM HEPES (Sigma-Aldrich), (pH 7.4). Input resistance was measured in the ‘whole cell’ configuration, and only cells that had an input resistance ≥ 0.5 GΩ were used for experiments. Cell capacitance and break-in resting membrane potential was also measured for each cell recorded. Data for current steps recordings was captured using a Multiclamp 700B amplifier controlled by pCLAMP (version 10.7.0.3), via an analogue-to-digital converter (Digidata 1440A, Molecular Devices). Whole cell recordings measuring synaptic drive to motoneurons and spontaneous motoneurons activity were conducted using an amplifier (Axopatch 1D) controlled by pCLAMP (V.10.3.1.5) by an analog-to-digital converter (Digidata 1322A, Molecular Devices. Trace/s were sampled at 20 kHz and filtered online at 10 kHz.

Current step recordings

Mecamylamine (200 µM in saline) was used to block postsynaptic nACh receptors in order to synaptically isolate neurons for experiments designed to measure intrinsic motoneuron excitability. Once patched, neurons were brought to a membrane potential of -60 mV using current injection. Each recording consisted of 20 x 4 pA (500 ms) current steps, including an initial negative step, giving a range of -4 to +72 pA. Number of spikes were counted and plotted against injected current. Both cell capacitance and input resistance were compared between conditions to ensure that any observed differences in excitability were not due to differences in either cell size or resistance.

Synaptic drive to motoneurons

To measure synaptic drive between A27h premotor interneurons and aCC, A27h neurons were identified via Chronos-mVenus reporter (as above), allowing corresponding aCC motoneurons to be patched. Recordings were conducted in current-clamp, with current injected into aCC sufficient to evoke a mean spike frequency of ∼10-20 Hz, prior to optogenetic stimulation. Recordings consisted of: individual sweeps, beginning with 1 s LED off, followed by 1 s LED on, and 1 s LED off (repeated 5-8 times per cell). Change in AP frequency (%) was calculated for each recording (per sweep) comparing AP frequency before and during stimulation, which was then averaged for each cell.

Spontaneous rhythmic currents (SRCs) were recorded for 180 s with aCC voltage clamped at -60 mV. Amplitude, duration, and frequency were measured for the first 15 events of each trace using Clampfit’s event detection, threshold search, function. The first 15 events with a stable baseline, an amplitude of >300 pA and no double peaks were accepted for analysis. Amplitude was measured as the change from baseline immediately prior to event initiation to the peak current and normalised for cell capacitance, providing a measure of current density. SRC duration was measured as the width of an event at 100 pA amplitude.

Spontaneous motoneuron recordings

Motoneuron spontaneous activity was recorded using an adapted whole cell current clamp protocol, in I=0, to measure passive action potential bursting properties. In Clampfit (version 10.3.1.5), 180 s recordings, were analysed to detect action potential ‘spike times’. These were imported into MATLAB and analysed using a custom script (developed by Mituzaite et al., 2021). An action potential burst was considered as an event consisting of at least 3 spikes occurring within 100 ms and ending when no spike was detected for 100 ms.

Crawling Assay

Crawling behaviour was recorded at the mid-third instar stage, 72 hours after larval hatching (ALH). Larvae were rinsed in water and placed inside a 24cm x 24cm crawling arena with a base of 5mm thick 0.8% agarose gel (Bacto Agar), situated inside an incubator. Temperature was maintained at 25±0.5°C, reported via a temperature probe in the agar medium. Humidity was kept constant. Larval crawling was recorded using a frustrated total internal reflection-based imaging method (FIM) in conjunction with the tracking software FIMTrack (Risse et al., 2017, 2013) using a Basler acA2040-180km CMOS camera, fitted with a 16mm KOWA-IJM3sHC.SW-VIS-NIR lens, acquisition controlled by Pylon software (Basler) and Streampix (v.6) software (NorPix). Larvae were recorded for 20 minutes at five frames per second.

Recordings were split into four 5-minute sections with the first five minutes of acclimatisation discarded. The remaining three 5-minute sections were used to analyse crawling speed, choosing crawling periods of uninterrupted forward crawling, devoid of pauses, turning or collision events. Each larva was sampled up to once per 5-minute section. Crawling speed was calculated using FIMTrack software.

Statistics

Statistical analyses were done using GraphPad Prism Software (Version 10.1.2). Datasets were tested for normal distribution with the Shapiro-Wilk Test. Normally distributed data were then tested with students t-test (for pairwise comparison). Normally distributed analysis for more than two groups was done with a one-way ANOVA and post hoc tested with a Tukey multiple comparison test. Non-normally distributed data sets of two groups were tested with Mann-Whitney U Test (pairwise comparison) and datasets with more than two groups were tested with a Kruskal Wallis ANOVA and post hoc tested with a Dunns multiple post hoc comparison test. For whole cell electrophysiology experiments in isolated neurons, linear regression analysis was used to compare the intercepts between conditions. This analysis was restricted to linear sections of data, represented by inset graphs. For all datasets mean and standard error of mean (SEM) are shown. Significance levels were * p<0.05; ** p<0.01; *** p<0.001; **** p<0.0001

eLife Assessment

This is an important study of critical period plasticity, focused on temperature manipulations, and how different parts of the Drosophila larval motor circuit adapt or maladapt. The work convincingly demonstrates that components of the motor network respond in distinct ways to the heat shock, and the combination of functional, structural, and electrophysiological approaches makes the study of significant interest. The work points to central interneurons as primary drivers of maladaptive changes, while motoneurons and neuromuscular junctions show compensatory or homeostatic adjustments. The study is methodologically rigorous, contributing significant insights into critical period biology using a tractable invertebrate model.

Reviewer #1 (Public review):

Summary:

The authors examine the impact of heat stress during an embryonic CP in Drosophila, focusing on the larval locomotor network. They show that elevated temperature increases neuronal activity and, when applied during the CP, results in long-term instability of the network, which manifests in prolonged seizure recovery times. At the neuromuscular junction, substantial structural changes occur, including terminal overgrowth and altered receptor composition, yet synaptic transmission remains preserved due to homeostatic regulation. Motoneurons display reduced excitability but receive increased synaptic input from premotor interneurons. These findings suggest that maladaptive instability originates within the central circuitry rather than at the neuromuscular junction, where changes seem to be homeostatically compensated. The study concludes that different network components exhibit distinct and hierarchical responses to CP perturbations, with premotor interneurons setting the tone for downstream adjustments in motoneurons.

Strengths:

The work takes advantage of the unique accessibility of the Drosophila system. A major strength of the study is the integration of structural, physiological, and behavioral analyses, which allows the authors to draw a comprehensive picture of how CP perturbations shape the locomotor network. The choice of an ecologically relevant stimulus (heat stress) is particularly convincing, as it links experimental manipulations more closely to natural environmental conditions. The experiments are carefully designed, and the results are robust and consistent with previous findings in the field, while also extending them in new directions.

Weaknesses:

The study leaves some uncertainty regarding the experimental design and interpretation. The change from short to prolonged heat shock manipulations raises the possibility that the effects observed may not be confined to the critical period alone - this could be experimentally addressed or simply rephrased in the text. In addition, the maladaptive (seizure recovery) and adaptive/homeostatic phenotypes are not always clearly distinguished or highlighted, which makes it harder to appreciate how the different levels of the network plasticity fit together into a single mechanistic framework.

Reviewer #2 (Public review):

Summary:

This manuscript presents a thoughtful and well-executed study of critical period plasticity in the Drosophila larval motor circuit. The authors examined how transient heat, 32 {degree sign}C, during the embryonic stage, altered network properties, showing that premotor interneurons A27h increase excitatory drive onto motoneurons, which respond with a reduction in excitability. At the NMJ, synaptic terminals expand and GluRIIA distribution shifts, yet synaptic transmission remains largely unaffected. Despite these local compensations, the treated larvae display slower crawling and prolonged recovery from seizures, indicating that the network is functionally compromised.

Strengths:

(1) One of the major strengths of this study is the elegant dissection of a defined circuit, tracking changes from premotor interneurons through motoneurons to the NMJ. The multimodal approach provides a comprehensive view of how connected elements respond to CP perturbations.

(2) An interesting finding is that NMJ morphology changes dramatically without corresponding deficits in synaptic transmission, challenging the common assumption that larger boutons necessarily indicate stronger synapses.

(3) Another intriguing result is that even with two layers of homeostatic compensation, locomotor behavior is still impaired, highlighting the limits of compensation and underscoring the critical role of CP timing.

(4) Beyond these scientific insights, the study benefits from a well-defined, tractable system and simple experimental manipulations, which together make the results highly interpretable and reproducible.

Weaknesses:

There are a few areas where the manuscript could be strengthened.

(1) Although A27h premotor neurons are well characterized, the claim that they are the causal driver of downstream changes would be strengthened by additional experiments or a clearer discussion of the temporal hierarchy.

(2) While 32 {degree sign}C heat stress is presented as ecologically relevant, it produces maladaptive behavioral outcomes, raising questions about the ecological and mechanistic interpretation of the model. In particular, most experiments, with the exception of Figure 1, used prolonged (24h) heat treatments, which could introduce developmental effects beyond the CP itself. Comparing shorter and longer heat exposures would help clarify the specificity of the CP response.

(3) While there are schematics for experimental procedures, a circuit diagram tracing information flow and indicating where structural and functional changes occur would help readers better understand the findings.

(4) Finally, the main paradox of the study, that robust homeostatic compensations occur yet behavior remains impaired, could be explored in more depth in the Discussion.

Reviewer #3 (Public review):

Summary:

During development, neural circuits undergo brief windows of heightened neuronal plasticity (e.g., critical periods) that are thought to set the lifelong functional properties of underlying circuits. These authors, in addition to others within the Drosophila community, previously characterized a critical period in late fly embryonic development, during which alterations to neuronal activity impact late-stage larval crawling behavior. In the current study, the authors use an ethologically-relevant activation paradigm (increased temperature) to boost motor activity during embryogenesis, followed by a series of electrophysiology and imaging-based experiments to explore how 3 distinct levels of the circuit remodel in response to increases in embryonic motor activity. Specifically, they find that each level of the circuit responds differently, with increased excitatory drive from excitatory pre-motor neurons, reduced excitability in motor neurons, and no physiological changes at the NMJ despite dramatic morphological differences. Together, these data suggest that early life experience in the motor neuron drives compensatory changes at each level of the circuit to stabilize overall network output.

Strengths:

The study was well-written, and the data presented were clear and an important contribution to the field.

Weaknesses:

The sample sizes and what they referred to throughout the distinct studies were unclear. In the legends, the authors should clearly state for each experiment N=X, and if N refers to an NMJ, for example, instead of an individual animal, they should state N=X NMJs per N=X animals. This will help readers better understand the statistical impact of the study.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment:

This is an important study of critical period plasticity, focused on temperature manipulations, and how different parts of the Drosophila larval motor circuit adapt or maladapt. The work convincingly demonstrates that components of the motor network respond in distinct ways to the heat shock, and the combination of functional, structural, and electrophysiological approaches makes the study of significant interest. The work points to central interneurons as primary drivers of maladaptive changes, while motoneurons and neuromuscular junctions show compensatory or homeostatic adjustments. The study is methodologically rigorous, contributing important insights into critical period biology using a tractable invertebrate model.

We thank the reviewers for their thoughtful critique and suggestions. We agree with these and, where possible, we have attempted to address these, improving this study. As outlined below, we have revised the manuscript substantively and included additional data and figures.

Public Reviews:

Reviewer #1 (Public review):

Summary:

The authors examine the impact of heat stress during an embryonic CP in Drosophila, focusing on the larval locomotor network. They show that elevated temperature increases neuronal activity and, when applied during the CP, results in long-term instability of the network, which manifests in prolonged seizure recovery times. At the neuromuscular junction, substantial structural changes occur, including terminal overgrowth and altered receptor composition, yet synaptic transmission remains preserved due to homeostatic regulation. Motoneurons display reduced excitability but receive increased synaptic input from premotor interneurons. These findings suggest that maladaptive instability originates within the central circuitry rather than at the neuromuscular junction, where changes seem to be homeostatically compensated. The study concludes that different network components exhibit distinct and hierarchical responses to CP perturbations, with premotor interneurons setting the tone for downstream adjustments in motoneurons.

Strengths:

The work takes advantage of the unique accessibility of the Drosophila system. A major strength of the study is the integration of structural, physiological, and behavioral analyses, which allows the authors to draw a comprehensive picture of how CP perturbations shape the locomotor network. The choice of an ecologically relevant stimulus (heat stress) is particularly convincing, as it links experimental manipulations more closely to natural environmental conditions. The experiments are carefully designed, and the results are robust and consistent with previous findings in the field, while also extending them in new directions.

Weaknesses:

The study leaves some uncertainty regarding the experimental design and interpretation. The change from short to prolonged heat shock manipulations raises the possibility that the effects observed may not be confined to the critical period alone - this could be experimentally addressed or simply rephrased in the text.

We agree that clarity about the experimental paradigm is important and have addressed this as suggested, within text, figures and figure legends: the duration of embryo exposure to 32°C heat stress is now unambiguously stated and each figure has a graphical illustrations of the heat stress paradigm. For example, experiments represented in Figures 1, 3 (new data) and 8 (new data) used short, defined periods of a few hours of heat stress, aimed to identify specific windows of development that are sensitive to 32°C heat stress. These also show that behavioural changes result from heat stress experienced during the specific 2-hour window that defines the critical period of the developing central locomotor circuitry, namely from 17-19 hours after egg laying - previously identified by Giachello & Baines (2015). Longer exposure to 32°C heat stress during embryogenesis result in the same phenotypes when this 2-hour window is included, causing the same level of reduced larval crawling speed and lowered network stability, which manifests in increased seizure recovery times. This is as one might expect from a critical period of nervous system development.

Where the neuromuscular junction is concerned, where we identified embryonic heat stress causing phenotypes that are evident at late larval stages, a more complex model has emerged. Following suggestions from both reviewers to explore shorter heat stress exposures during embryogenesis, additional experiments (see Figure 3) we identified what might be a critical period for the body wall muscles. This is an earlier window of development, within 13-16 hours after egg laying, which is sensitive to heat stress in terms of the levels of the GluRIIA glutamate receptor subunit that will be expressed in the late larva. This developmental period is characterised by muscles acquiring their electrical properties (Broadie & Bate, 1993), i.e. comparable to the central locomotor network transitioning through its critical period at the time that it becomes active. The neuromuscular junction is composed of both presynaptic motoneurons and postsynaptic muscles, and therefore this composite structure is subject to multiple, sequential critical periods. The characterisation of changes to neuromuscular junction synaptic physiology was carried out using heat stress throughout most of embryogenesis (Figure 4). We think this appropriate from the perspectives of having included all relevant critical periods (muscle and CNS) to explore how this composite structure responds to environmental heat stress; also based on our observations that for each critical period phenotypes are defined by the experience during the critical period and not exacerbated by prolonged heat stress either side.

In addition, the maladaptive (seizure recovery) and adaptive/homeostatic phenotypes are not always clearly distinguished or highlighted, which makes it harder to appreciate how the different levels of the network plasticity fit together into a single mechanistic framework.

Following the suggestion, we have tried to clarify the mechanistic framework in a new figure that aims to summarise the model in Figure 9.

The question of whether phenotypes that result from an embryonic heat stress manipulation are adaptive or maladaptive is difficult to resolve. This is partly due to the nature of critical periods, since perturbations during these developmental windows can cause significant, long-lasting maladaptations that are challenging to reconcile from a perspective of adaptive plasticity. Secondly, in light of the nature of this animal, which has evolved a particularly rapid development and large brood sizes, any deviation from the evolved optimum developmental temperature of 25°C could constitute a reduction in fitness. We interpret the phenotypes we see along those lines: network instability that results from critical-period perturbations is a manifestation of a sub-optimally tuned network, as is a reduction in larval crawling speed.

Reviewer #2 (Public review):

Summary:

This manuscript presents a thoughtful and well-executed study of critical period plasticity in the Drosophila larval motor circuit. The authors examined how transient heat, 32 {degree sign}C, during the embryonic stage, altered network properties, showing that premotor interneurons A27h increase excitatory drive onto motoneurons, which respond with a reduction in excitability. At the NMJ, synaptic terminals expand and GluRIIA distribution shifts, yet synaptic transmission remains largely unaffected. Despite these local compensations, the treated larvae display slower crawling and prolonged recovery from seizures, indicating that the network is functionally compromised.

Strengths:

(1) One of the major strengths of this study is the elegant dissection of a defined circuit, tracking changes from premotor interneurons through motoneurons to the NMJ. The multimodal approach provides a comprehensive view of how connected elements respond to CP perturbations.

(2) An interesting finding is that NMJ morphology changes dramatically without corresponding deficits in synaptic transmission, challenging the common assumption that larger boutons necessarily indicate stronger synapses.

(3) Another intriguing result is that even with two layers of homeostatic compensation, locomotor behavior is still impaired, highlighting the limits of compensation and underscoring the critical role of CP timing.

(4) Beyond these scientific insights, the study benefits from a well-defined, tractable system and simple experimental manipulations, which together make the results highly interpretable and reproducible.

Weaknesses:

There are a few areas where the manuscript could be strengthened.

(1) Although A27h premotor neurons are well characterized, the claim that they are the causal driver of downstream changes would be strengthened by additional experiments or a clearer discussion of the temporal hierarchy.

We have tried to clarify the model of the temporal hierarchy (new Figure 9). This is a model and as such will hopefully help us collectively to think about this system and underlying processes, while also inviting this perspective to be challenged. The model we propose is compatible with our observations, namely that the premotor circuitry might change in response to a critical period heat stress (e.g. increasing their synaptic drive onto motoneurons), followed by homeostatic adjustment by the postsynaptic motoneurons (e.g. by reduction of their excitability), thus serving to maintain overall normal motoneuron firing patterns (see Figure 6).

However, synaptic communication is commonly regulated in both antero- and retrograde directions. Therefore, while compatible with the observations we have made, bi-directional information flow could also be instructive during the CNS critical period.

(2) While 32 {degree sign}C heat stress is presented as ecologically relevant, it produces maladaptive behavioral outcomes, raising questions about the ecological and mechanistic interpretation of the model. In particular, most experiments, with the exception of Figure 1, used prolonged (24h) heat treatments, which could introduce developmental effects beyond the CP itself. Comparing shorter and longer heat exposures would help clarify the specificity of the CP response.

We agree. For a detailed response on this point, please response to Reviewer #1 above.

(3) While there are schematics for experimental procedures, a circuit diagram tracing information flow and indicating where structural and functional changes occur would help readers better understand the findings.

We have created Figure 9 as a working model.

(4) Finally, the main paradox of the study, that robust homeostatic compensations occur yet behavior remains impaired, could be explored in more depth in the Discussion.

We have tried to address this in the discussion.

Reviewer #3 (Public review):

Summary:

During development, neural circuits undergo brief windows of heightened neuronal plasticity (e.g., critical periods) that are thought to set the lifelong functional properties of underlying circuits. These authors, in addition to others within the Drosophila community, previously characterized a critical period in late fly embryonic development, during which alterations to neuronal activity impact late-stage larval crawling behavior. In the current study, the authors use an ethologically-relevant activation paradigm (increased temperature) to boost motor activity during embryogenesis, followed by a series of electrophysiology and imaging-based experiments to explore how 3 distinct levels of the circuit remodel in response to increases in embryonic motor activity. Specifically, they find that each level of the circuit responds differently, with increased excitatory drive from excitatory pre-motor neurons, reduced excitability in motor neurons, and no physiological changes at the NMJ despite dramatic morphological differences. Together, these data suggest that early life experience in the motor neuron drives compensatory changes at each level of the circuit to stabilize overall network output.

Strengths:

The study was well-written, and the data presented were clear and an important contribution to the field.

Weaknesses:

The sample sizes and what they referred to throughout the distinct studies were unclear. In the legends, the authors should clearly state for each experiment N=X, and if N refers to an NMJ, for example, instead of an individual animal, they should state N=X NMJs per N=X animals. This will help readers better understand the statistical impact of the study.

This is a good point. For the majority, each data point is derived from a unique specimen, unless explicitly stated otherwise, for NMJ size on muscle DA1 (Figure 3) and for larval crawling data, where each larva was measured up to three time, once per unique 5-minute crawling interval.

Recommendations for the authors:

Reviewing Editor Comments:

In addition to revising the text and making interpretive changes as suggested by the reviewers, we invite you to consider the following:

(1) Either rephrase the conclusions on the role of the 2h CP and discuss the effects of temperature during embryonic development. Alternatively, to validate the idea of a longer CP window, directly compare the results of a few key experiments using the 2h and 24h heat treatment.

We have addressed this within the text, as suggested. In the text and figure legends, we have made clear distinctions between exposure to 32°C heat stress during most of embryogenesis vs a specific developmental window of a few hours. In figures, we have provided diagrams that graphically illustrate the period of heat stress exposure.

Our ability to experimentally test differences between precise vs broader heat stress periods during embryonic development have been constrained due to the departure of scientists, who were able to carry out electrophysiological recordings (as also explained below). As the next best alternative, we focus on imaging and behavioural analyses. These demonstrated that the developing body wall muscles are sensitive to heat stress during an earlier phase, from 13-16 hours after egg laying, when the body wall muscles become electrically active. It precedes the critical period of the central locomotor circuitry (17-19 hours after egg laying), when neurons in the CNS become electrically and synaptically active (16 hours after egg laying). We think this an exciting additional insight, demonstrating sequential critical periods as different parts of the locomotor network become active: first the body wall muscles, followed by the central circuitry.

(2) Clarifying the homeostatic responses and shedding light on how they engage with the maladaptive changes described would benefit the study. Furthermore, adding more information about the anatomical and structural changes and how they relate to the intrinsic and synaptic changes would also benefit the study.

We have tried to address this within the text and with a summary diagram (Figure 8), as suggested.

Reviewer #1 (Recommendations for the authors):

(1) It remains unclear whether the authors want to conclude that reduced network stability is not due to changes at the motoneuron level, but rather at the premotor level. Although this idea is mentioned in the results and discussion, it does not appear in the abstract or introduction, which leaves the different findings disconnected. Clarifying and highlighting this conclusion throughout the manuscript would strengthen the narrative.

We have added additional experiments and changed the manuscript to address this point. These showed that there are distinct phases of embryonic development during which heat stress causes changes to NMJ structure vs to larval behaviour (seizure recovery times/network stability and crawling speed) - additional data in suppl. Fig. 2 and Fig. 7). In the Drosophila embryo, the body wall muscles develop and acquire their electrical properties before central neurons do, and these phases correlate with sensitivity to heat stress.

(2) In the results section related to Figure 1, the logical link between the CP protocol and the functional assessment of the locomotor network at different temperatures is not sufficiently explained. It is not clear what this assessment is meant to test or demonstrate. A more explicit statement of the rationale and correction of what seems to be a typographical error in the final sentence of the paragraph would help to clarify the authors' intent.

We have tried to rectify this by changes in the manuscript and to Figure 1, to make the sequence of panels more intuitive.

(3) In the second results section, the experimental strategy shifts from using a short 2-hour heat shock to a 24-hour manipulation. The reasoning - that short manipulations in different windows yield no phenotype - is understandable, but a 24-hour perturbation may have broader consequences beyond the CP, simply by virtue of its longer duration. Moreover, 24h is roughly the duration of embryonic development at 25C. When at 32C, embryos should develop faster; therefore, is the 24h heat shock extending to L1?

Yes, the 24 hour heat stress extends into the first few hours of the L1 larval stage.

In order to validate the use of a longer window, the author should show how it affects the developmental time. Moreover, one should test that a few important observations remain the same with 2h and 24h heat perturbation. Alternatively, one cannot conclude that the phenotype is due to the rather narrow previously defined CP rather than to other effects associated with the overall embryonic developmental time and coordination. This would not make the results less interesting, but it would be important to assess whether the effects can be solely attributed to the 2h CP.

We have compared the impact of heat stress experience during the majority of embryogenesis, including the CP that had been defined for the central locomotor network (17-19 hours after egg laying) with shorter heat stress manipulations during consecutive phases of embryogenesis until larval hatching. As outlined above in response to point (1) by the Reviewing Editor, reduced stability of the central network and associated reduction in larval crawling occurs when heat stress is experienced during the CP of the central locomotor network (17-19 hours after egg laying). Prolonged heat stress experience for 24 hours leads to indistinguishable outcomes, as long as this 2-hour CP window is included (see Fig. 1 and Fig. 8).

However, this suggestion by Reviewer #1 led us to identify a second CP for the body wall muscles (see Fig. 3). NMJ overgrowth and changes to the postsynaptic glutamate receptor composition result from earlier heat stress experiences, and those are comparable to the effects caused by 24-hour heat stress exposure when this earlier muscle CP is included.

Therefore, NMJ development is affected by consecutive CPs, an earlier one linked to body wall muscle development, followed by a later one that impacts the presynaptic motoneurons and their upstream circuitry. Nevertheless, the larval NMJ and behavioural phenotypes that we have identified appear to result from sensitivity to heat stress during these respective CP windows, with no clear evidence of cumulative effects on these phenotypes resulting from longer heat stress exposure during embryogenesis.

(4) In session 3, the authors note that GluRIIA reductions were most pronounced in proximal regions of the NMJ. However, this is not explicitly quantified in the figures or methods. Including such quantification, or clarifying where it can be found, would make this observation more convincing.

We have analysed anti-GluRIIA signal intensities in proximal vs distal boutons, comparing different ROI selection processes (e.g. thresholding to a full NMJ/anti-HRP mask and to an anti-GluRIIB mask, which is more selective to postsynaptic sites). Analysis of multiple data sets did not show statistical significance, but instead confirmed that comparable reductions in anti-GluRIIA signal manifest in both proximal and distal boutons, following an embryonic 32C heat stress, relative to controls. We have therefore removed relevant speculative statements.

(5) In session 4, the authors conclude that motoneurons undergo a decrease in excitability to adjust to greater premotor drive. Is there anatomical evidence for this, such as an increase in input synapses?

We previously quantified change in excitatory presynaptic synaptic contact number onto aCC motoneuron dendrites in third instar larvae following an embryonic pharmacological activity manipulation: overexcitation of the developing network following introduction of PTX via feeding to gravid females*. No significant structural changes were seen. Although this is a different manipulation of the developing network, all our data to date suggest that heat stress manipulations during the embryonic critical period signal via the same pathways, at least in part due to temperature increases leading to activity increases. Because such a quantification is technically challenging and extremely time-consuming due to the low level of marking individual motoneurons, we did not think it informative or in scope for this project.

*See Figure 5 in this publication: Hunter I, Coulson B, Pettini T, Davies JJ, Parkin J, Landgraf M, Baines RA. Balance of activity during a critical period tunes a developing network. Elife. 2024 Jan 9;12:RP91599. doi: 10.7554/eLife.91599. PMID: 38193543; PMCID: PMC10945558.

The interpretation of the optogenetic experiments would also benefit from clarification. If motoneurons are less excitable yet receive more drive, one might expect no net change, rather than the differences observed. Alternatively, could the excitability of the premotor neuron itself have changed, either intrinsically or in relation to Chronos expression? Measuring premotor activity directly during optogenetic activation could help to resolve this ambiguity.

These are good suggestions. Yes, we think that motoneuron excitability has changed as a result of heat stress - see paper submitted in parallel and published since: Sobrido-Cameán et al., 2025, PLoS Biology. Unfortunately, the team members, who could have carried out this type of analysis had moved on by submission of the manuscript. Therefore, we were no able to experimentally pursue these questions further.

(6) In session 6, the authors report slower propagation of premotor activity waves after CP heat stress, but the logic of the experiment is not sufficiently explained. How does this finding relate to the enhanced premotor drive described earlier? Only timing is quantified; information about amplitude and wave dynamics would strengthen the interpretation. These results could also be discussed in relation to Figure 1B, where acute heat stress increased motoneuron activity. One interesting possibility could be that CP manipulations might adaptively prepare the larva to function at different temperatures. Experiments testing wave propagation at 32 {degree sign}C (Figure 6) or, conversely, motoneuron activity after CP manipulations (Figure 1B paradigm) would provide valuable evidence for such an adaptive role.

As per above, unfortunately, the team member who could have carried out this type of analysis had moved on by submission of the manuscript. However, we have tried to address the question of whether there is an adaptive element to the adjustments that result from embryonic heat stress experience. Specifically, we carried out behavioural tests on how larvae respond with changes in crawling speed to acute changes in ambient temperature (new Fig. 7).

We interpret our findings as follows: that heat stress during embryonic development leads to sub-optimal outcomes with regard to network stability as well as default and maximum crawling speed. The precise causes for this will be difficult to unpick. Behaviourally, when challenging larvae with an acute change in ambient temperature, we saw that slow crawling larvae do respond comparatively normally to a relative increase in ambient temperature by speeding up (in effect an escape response). This demonstrates that embryonic heat stress causes a change in the default crawling speed, while principally maintaining behavioural responses to changes in ambient temperature. It appears that animals that had experienced heat stress during embryonic development, by adjusting their default speed downward, maintain a dynamic response range into a higher temperature range than controls (35C vs 29C, respectively). Potentially, this could be an adaptive outcome to living at higher temperatures, though such an interpretation would require a body of work.

Nevertheless, every aspect we have assayed suggests that heat stress experience during the CP leads to sub-optimal outcomes: of network instability, slower default and slower maximal crawling speeds.

Minor points:

(1) Abstract: "has suboptimal outcomes;" should be corrected to "has suboptimal outcomes,".

Corrected

(2) Abstract: "we find that transient embryonic..." would improve readability with a capitalized "We".

We are unsure about the sentence this refers to. If this sentence, then we suggest that this could remain as was.

"Within the central nervous system, we find transient embryonic CP perturbation leads to increased synaptic drive from premotor interneurons to motoneurons..."

The intention here is to differentiate between changes within the CNS vs at the NMJ.

(3) Abstract: The sentence "Present the larva ... as an experimental model system..." overstates novelty, as the system has already been established in prior work. Instead, this study could highlight temperature manipulation as an ecologically relevant way to probe CPs.

We have adopted this suggestion.

Reviewer #2 (Recommendations for the authors):

I would recommend:

(1) Perform additional experimental support and dissuasions for the causal role of the premotor neuron and the network disability.

Unfortunately, it has not been possible to carry out additional e-phys experimental work due to key people having moved on and now unable to carry out such experiments, and no replacements in sight to do so. Instead, we have focused on other work that we could do, namely to test the effect of different heat stress windows during embryonic development on GluRIIA vs GluRIIB expression at postsynaptic sites. This shows that indeed the effect seen following a 24-hour heat stress is replicated by a much shorter window of heat stress. For NMJs GluRII composition the critical period is different from the critical period of the CNS. This replicates the different developmental timings of maturation: the body wall muscles express ion channels and attain their electrical properties several hours before central neurons. We have provided these additional data as a new supplementary figure to Fig. 2. We have changed the main text to note the caveat of longer heat stress manipulations potentially leading to additional or more exacerbated phenotypes.

(2) I would suggest that the authors expand the discussion on why two layers of homeostatic adjustment fail to preserve behavior. Is this simply a limit of plasticity?

This is a difficult aspect to address well, beyond the purely speculative and potentially confusing. We would like to suggest that to do so requires a basic understanding on what pressures neurons/networks respond to (heat-caused over-activation and/or metabolic); and from that perspective to gauge how they adjust to those pressures, i.e. what the adjustments are trying to "achieve". This in turn should inform on whether such adjustments are homeostatic or anti-homeostatic in nature, and whether there are limits beyond which we consider a system "breaking".

(3) 32C heat was described as "ecologically relevant". However, it produces maladaptive outcomes. The author should consider reframing it as a stressor that reveals CP sensitivity, rather than an adaptive signal.

This is a good suggestion, and we have implemented changes accordingly.

(4) It is important to distinguish the effects of transient (2h) vs prolonged heat exposure to confirm the manipulation targeted CP specifically.

We have tried to make these distinctions clearer within text and figure legends. As per response to (1) above, we generated and analysed additional data to show that changes in GluRIIA are induced during a defined shorter developmental time window, not exacerbated by prolonged heat stress exposure during embryogenesis.

(5) I would definitely recommend adding a diagram tracking information flow and showing where structural and functional changes occur.

This is a helpful suggestion, which we have tried to implement with a new figure (Fig. 8).

Overall, this is a strong and well-written paper that produced some unexpected results, and added a solid model circuit to study CP plasticity at the circuit level.

Reviewer #3 (Recommendations for the authors):

I identified one typo: "activity manipulations during the embryonic CP are artificial, We asked to what extent" The "W" of "We" should be lower case.

Now corrected.

Version 2

Abstract

As developing neural circuits become functional, they undergo a phase of heightened plasticity that facilitates network tuning in response to intrinsic and/or extrinsic stimuli. These developmental windows are termed critical periods (CPs), because perturbations during, but not outside, the CP can lead to lasting changes, such as the formation of sub-optimal or unstable networks. How separate, but connected elements, within a network might respond differently to a CP perturbation is not well understood. To study this, we used the locomotor network of the Drosophila larva as a model and heat stress as a CP stimulus that has ecological relevance. When embryos experienced heat stress the subsequent development of their locomotor network is changed, creating larvae with reduced crawling speed and decreased network stability. Developing body wall muscles and central neurons are sensitive to heat stress perturbations during distinct, consecutive phases of embryogenesis. Within the CNS, transient embryonic CP perturbation leads to increased synaptic drive from premotor interneurons to motoneurons, which in turn adopt reduced excitability. In contrast, the peripheral neuromuscular junction maintains normal synaptic transmission, despite significant structural changes. Thus, connected elements respond differentially to a CP perturbation, suggesting a sequence, or hierarchy, of network adjustment during the CP.

Introduction

During nervous system development, critical periods (CPs) are transient windows of heightened plasticity, which are fundamental to the emergence of appropriate network function. Because of this, networks are particularly vulnerable to activity perturbations during a CP, which can result in lasting structural and/or functional changes, while comparable perturbations outside the CP are less impactful (Coulson et al., 2022; Reh et al., 2020). The functions and concepts of CPs have traditionally been investigated in the visual system of mammals (Hensch, 2005; Hensch and Quinlan, 2018; Hubel and Wiesel, 1970) and imprinting of birds (Hess, 1975; Horn, 2004). More recently, CPs have also been identified and investigated across a range of species and neuronal ensembles, from networks underlying the development of speech (Harrison et al., 2005) to motor behaviour of rats (Soiza-Reilly and Azcurra, 2009), fish (Hageter et al., 2023) and insects (Dombrovski and Condron, 2021; Giachello and Baines, 2015). The presence of a CP in developing networks is therefore increasingly considered a universal phenomenon. The precise changes within networks and associated mechanisms, as they occur during normal development, as well as those resulting from perturbations during a CP, are not well understood. This is in part due to the complexity of many of the experimental model systems, notably mammalian cortical systems that have been the predominant model for studying CPs.

To overcome this complexity, we have utilized the comparatively simpler nervous system of the fruit fly larva, Drosophila melanogaster, with which to study CPs. Though an insect, Drosophila displays CPs with features that suggest commonalities to mammalian CPs (for review see: Coulson et al., 2022). Currently, the best characterised CP during Drosophila development is one that occurs during embryogenesis, linked to the emergence of normal locomotor network output (Carreira-Rosario et al., 2021; Corke et al., 2025; Coulson et al., 2022; Crisp et al., 2008, 2011; Ackerman et al., 2021; Fushiki et al., 2013; Giachello et al., 2021; Giachello and Baines, 2015; Hunter et al., 2024; Sobrido-Cameán et al., 2025; Williamson et al., 2021; Zeng et al., 2021). This experimental system benefits from the comprehensively annotated connectome of the Drosophila larval nervous system (Giachello et al., 2020; Heckman and Doe, 2022; Valdes-Aleman et al., 2021; Winding et al., 2023) as well as an unparalleled set of genetic tools for manipulating identified neurons with both spatial and temporal precision (Simpson, 2009). Combined with imaging and electrophysiological techniques, this model system enables investigation of cellular responses to CP manipulations at the level of identified cells and synapses (Carreira-Rosario et al., 2021; Coulson et al., 2022; Crisp et al., 2008, 2011; Fushiki et al., 2013; Giachello et al., 2021; Giachello and Baines, 2015; Hunter et al., 2024; Sobrido-Cameán et al., 2025; Williamson et al., 2021; Zeng et al., 2021).

Neuronal activity manipulations during the CP of Drosophila embryos have been shown to cause decreased network stability. This manifests in significantly longer times required for larvae to recover from a strong activity challenge, such as electric shock-induced seizures (Coulson et al., 2022; Giachello et al., 2021; Giachello and Baines, 2015; Hunter et al., 2024, 2021). The premise that balanced network activity during the CP is necessary for stable, resilient networks to form has been explicitly supported by findings from our recent studies (Giachello et al., 2021; Giachello and Baines, 2015; Hunter et al., 2024). By extension, rectifying activity manipulations during identified CPs could have therapeutic potential (Nardou et al., 2023).

While experimental perturbations of neuronal activity during development have been a useful methodological approach for studying CPs as a phenomenon and the associated biological mechanisms, we tested if temperature change, in the form of 32°C heat stress, might be an alternative perturbation stimulus, with ecological relevance. For example, most animals are not warm-blooded and in poikilothermic animals, such as Drosophila, changes in ambient temperature, daily as well as seasonal, lead to changes in behavioural activity, likely caused by changes in neuronal and network activity (Lee and Montell, 2013; Powsner, 1935). Studies from the crustacean stomatogastric nervous system showed that these minimal networks produce a characteristic, rhythmic firing pattern, which is robustly maintained across a range of external temperatures. Importantly, neuronal network properties were adapted relative to the long-term temperatures that crabs had experienced, resulting in significantly different upper “crash temperatures” at which networks ceased to function (Marder, 2023; Marder and Rue, 2021). In Drosophila, the temperature experienced by late embryos determines larval locomotor output, including crawling speed and social feeding behaviour (Williamson et al., 2021). Similarly, studies on the Drosophila adult visual and olfactory systems showed that neuronal wiring properties and associated behaviours of adults are determined by the temperature experienced during a second CP of Drosophila, which takes place during pupariation (Chodankar et al., 2020; Kiral et al., 2021; Züfle et al., 2025), a phase of extensive network remodelling (Baumann et al., 2024; Leier et al., 2025; Lowe et al., 2023; Nelson et al., 2024).

In this study, we used heat stress of 32°C as an ecologically relevant stimulus with which to probe the embryonic CP. To gain insights at the level of single, identified connected cells, we focused on the neuromuscular system of the Drosophila larva, and found that the components of this network respond differently to heat stress experienced during embryonic development. Specifically, we find that elevated temperature increases neuronal activity and, when experienced during the locomotor CP, causes long term reduced network stability, as seen for neuronal activity manipulations (Giachello et al., 2021; Giachello and Baines, 2015; Hunter et al., 2024). At the neuromuscular junction (NMJ), embryonic heat stress experience causes presynaptic terminal overgrowth and changes in postsynaptic neurotransmitter receptor composition, though synaptic transmission remains normal. In contrast, within the central circuitry, we find that embryonic heat stress leads to enhanced synaptic drive from premotor interneurons and, potentially as a response, reduced excitability of motoneurons. Overall, this study presents a first comprehensive analysis of how several interconnected elements of a defined locomotor network develop in response to CP perturbation by heat stress, an ecologically relevant stimulus.

Results

Temperature manipulation during the locomotor CP leads to an unstable crawling network

We had previously identified a CP of nervous system development in late embryogenesis, from 17 – 19 hrs after egg laying (AEL), equivalent to 80-90% of embryonic development (Giachello and Baines, 2015). Perturbation of network activity in late embryogenesis, during this CP, pharmacologically or optogenetically, leads to the formation of an unstable, seizure prone network, as revealed by acute activity challenges. For example, an electric shock administered to late larvae, which develop from CP-manipulated embryos, causes seizures in those larvae, with recovery times that are significantly longer compared to unmanipulated controls. These pharmacological and optogenetic large-scale activity manipulations, while useful to study the embryonic CP, are artificial. We therefore asked to what extent changes in environmental conditions, that embryos might normally encounter, could also impact on network development. As poikilothermic animals, Drosophila development strongly depends on ambient temperature (Powsner, 1935). Several studies have shown that temperature experiences during development alter neuronal wiring as well as behaviours in larvae and in adult flies (Kiral et al., 2021; Williamson et al., 2021; Züfle et al., 2025). Increases in ambient temperature can lead to increased activity in the locomotor network (Oswald et al., 2018; Sigrist et al., 2003; Zhong and Wu, 2004). We verified this by functional imaging in isolated CNS, measuring spontaneous calcium transients from motoneurons, via GCaMP8f selectively expressed in aCC motoneurons using R94G06-GAL4. Stepwise increases in temperature led to increases in the frequency of rhythmic bursts. In contrast, CNS kept at the control temperature of 25°C for the duration of functional imaging (600 seconds) tended to show a decrease in frequency over time (Fig. 1A). These experiments suggest that increases in ambient temperature, above 25°C, lead to concomitant increases in locomotor network activity.

Transient heat stress during the critical period leads to an unstable network.

A) Acute increase of ambient temperature leads to increased network activity, as measured with the fluorescent calcium indicator GCaMP8f expressed in segmentally repeated aCC motoneurons in isolated central nervous systems; while constant 25°C leads to reduced activity over time. Left: micrograph of an isolated ventral nerve cord with segmentally repeated aCC motoneurons expressing GCaMP8f. Centre: traces of GCaMP fluorescence peaks measured at different temperatures. Right: quantification of fluorescence peak frequency changes over time (control/sham -bottom trace) and in response to temperature increase (top trace). B) Heat stress, when experienced during the embryonic CP of locomotor network development, leads to reduced network stability. Left: Schematic representation of embryonic and larval development and the paradigm of temperature manipulation during embryogenesis. Embryos were kept at the control temperature of 25°C, either continuously (control) or moved to 32°C for 2 hours, either before (13-15 h, 15-17 h AEL), during (17-19 h AEL) or after (19-21 h AEL) the locomotor network CP, then returned to 25°C until the late larval stage, when subjected to seizure induction by electroshock. Right: quantification of seizure recovery times. n=30, data are shown with mean ± SEM, ANOVA, ****p < 0.00001, ‘ns’ indicates statistical non-significance.

Next, we tested the effect that a transient experience of 7°C above the evolved temperature preference of 25°C (Ashburner et al., 1983), namely 32°C, might have on nervous system development. We chose 32°C as this is known to be experienced by Drosophila melanogaster (Williamson et al., 2021). We transiently exposed embryos for two hours to 32°C heat stress either before, during, or after the previously characterised embryonic CP for locomotor network development (17 – 19 hrs AEL) (Giachello and Baines, 2015), then returned them to the control temperature of 25°C. After four days, at the end of larval development, we measured as a proxy for network stability the speed of recovery from electroshock-induced seizures. Seizure recovery was unchanged, at control levels, following transient heat stress before or after the embryonic CP. In contrast, animals that had experienced 32°C heat stress during the embryonic CP (i.e. between 17 – 19 hrs) had significantly prolonged recovery times, indicative of reduced network stability (Fig. 1B). This shows that when Drosophila embryos experience a 32°C heat stress during the embryonic CP of the locomotor network it results in altered developmental outcomes, notably the formation of sub-optimal, less stable networks, similar to the effect of network-wide activity perturbations during this developmental window (Hunter et al., 2024; Giachello et al., 2021; Giachello and Baines, 2015).

To investigate structural and functional changes that result from a CP heat stress perturbation we first focused on the well characterised, and easy to access, peripheral neuromuscular junction (NMJ). Specifically, we focused on the NMJ of dorsal acute muscle 1 (DA1), which is innervated by the aCC motoneuron, a cell that has been the focus of our previous studies (Fig. 2A) (Baines et al., 2001, 1999; Fujioka et al., 2003; Hunter et al., 2024; Landgraf et al., 2003; Giachello et al., 2021; Giachello and Baines, 2015 Sobrido-Cameán et al., 2025). Because temperature manipulations outside the CP did not lead to an unstable network (Fig. 1A), we adopted a simplified experimental paradigm of putting embryos from a 6h egg collection to 32°C for 24h. This produces outcomes comparable to heat stress exposure centred around the CP, while ensuring that all embryos go through their CP during heat stress irrespective of minor variations in developmental timing or due to retention of fertilised eggs. We find that heat stress during embryogenesis leads to a significant overgrowth of the aCC motoneuron terminal (composed of type Ib boutons) by the late larval stage. This increase in the number of boutons is accompanied by a corresponding increase in active zones, while overall active zone density is not altered (Fig. 2B). On the postsynaptic side, embryonic heat stress causes an altered receptor composition: a selective reduction of ion channels containing the larger conductance GluRIIA subunit. Across the aCC NMJ, this leads to a reduced ratio of the larger conductance GluRIIA versus the lower conductance GluRIIB subunit-containing glutamate receptors (Fig. 2C).

Morphological changes at the neuromuscular junction after critical period perturbation.

A) Schematic of temperature manipulation during embryogenesis and location of dorsal acute muscle 1 (DA1) within a fillet dissected larva. Embryos collected over a 6h period were incubated at 32°C for 24h to guarantee heat stress exposure during the locomotor network CP. B) This results in a change of subsequent NMJ development, which manifests as a significant overgrowth of aCC presynaptic terminals (green, visualised with anti-HRP-AlexaFluor488). The increase in bouton number and NMJ area is paralleled by concomitant increases in active zones (magenta), resulting in a normal active zone density. C) Postsynaptically, embryonic heat stress causes in a significant decrease of glutamate receptors containing the larger conductance GluRIIA subunit, while those with the GluRIIB subunit remain unaffected. Thus, the postsynaptic GluRII receptor composition is significantly changed following a transient embryonic heat stress, as reflected in the ratio of receptors containing GluRIIA vs GluRIIB. Data are shown with mean ± SEM, unpaired t-test, ***p < 0.0001, ****p < 0.00001, ‘ns’ indicates statistical non-significance.

In recognition of the potential caveat that prolonged heat stress during most of embryogenesis could lead to compound effects, we additionally carried out 32°C heat stress exposure of embryos during shorter time windows, followed by rearing at the control temperature of 25°C. We then measured postsynaptic GluRIIA vs GluRIIB levels at the aCC NMJ on muscle DA1 of late wandering third instar larvae. We found that the GluRIIA reductions caused by prolonged heat stress throughout most of embryogenesis is recapitulated by a shorter, 3h exposure to 32°C, though during a different developmental window, from 13-16 hrs AEL. Heat stress before or after this 13-16 hrs AEL developmental window had no effect on the composition of the postsynaptic glutamate receptors (Fig. 3). GluRIIB subunit levels, in contrast, remained unaffected by embryonic exposure to 32°C heat stress. The period of 13-16 hrs AEL is the developmental window during which the embryonic body wall muscles become electrically active and start to exhibit myogenic contractions (Broadie and Bate, 1993; Crisp et al., 2008). Thus, the maturation of the embryonic body wall muscles precedes the CP of the central locomotor circuitry (17-19 hrs AEL), when central neurons develop their electrical properties (Baines and Bate, 1998). To our surprise, we found that NMJ size is impacted by heat stress during an even earlier window of muscle development (9-12 hrs AEL), during which myoblast fusion occurs (Bate et al., 1993) (Fig. 3). Overall, we found that transient embryonic heat stress of 32°C leads to reproducible changes in NMJ development during subsequent larval stages. Our observations show that postsynaptic glutamate receptor composition is specified during the developmental window that is characterised by electrical maturation of the body wall muscles, preceding by several hours the equivalent maturation of the innervating motoneurons.

Heat stress pertubations during distinct periods of development differentially affect postsynaptic GluRII receptor composition vs NMJ size.

A) Schematic of temperature manipulations during embryogenesis. Embryos collected over one hour were incubated at 32°C for different phases of embryogenesis. B) Resultant changes to the development of the aCC NMJ on muscle DA1 were quantified at the late wandering third instar larval stage. Reduced expression of the GluRIIA subunit occurred following heat stress experienced during 13-16 hours AEL, comparable to GluRIIA changes caused by heat stress during all of embryogenesis, but not in response to heat stress experienced before or after this 3-hour window. GluRIIB levels are not affected by heat stress during any period or all of embryogenesis. NMJ size, in contrast, is affected by heat stress experienced during earlier phases of muscle development, from 9-13 hours AEL. Data are shown with mean ± SEM, ANOVA, *p < 0.01, **p < 0.001, ***p < 0.0001, ****p < 0.00001, ‘ns’ indicates statistical non-significance. Sample sizes: for GluRIIA and GluRIIB measurements, each data point represents a unique specimen. NMJ size was measured in the same specimens for up to four muscle DA1 NMJs in abdominal segments 3-5 sampled per larva.

Despite morphological changes, larval NMJ physiology is not altered

In view of the prominent changes in NMJ growth, morphology and postsynaptic glutamate receptor composition that result when heat stress (32°C) is experienced during embryonic development, we next assessed NMJ synaptic physiology. Because the NMJ is a composite structure that develops through interactions between the postsynaptic body wall muscle with presynaptic terminals from motoneurons, whose cell bodies are located within the CNS, we proceeded using an experimental paradigm of prolonged heat stress (32°C) during most of embryogenesis, so as to include CPs of both body wall muscles and the central locomotor circuitry. We used the two-electrode voltage clamp (TEVC) technique to assess muscle synaptic physiology, initially measuring single pulse transmission across a gradient of external calcium concentrations. Despite the marked structural changes in heat stress manipulated animals, single pulse transmission at the NMJ was not significantly different between embryonically manipulated (32°C experience during embryogenesis) and controls that had been reared at 25°C throughout (Fig. 4 A&B). Quantification of the number of release sites utilised per action potential and quantal size (Jusyte et al., 2023) also showed no statistically significant difference between manipulated and control specimens (Fig. 4C). We quantified the size of the readily releasable pool (RRP) of synaptic vesicles by recording responses to a high frequency train of action potentials (Fig. 4D), then plotting the cumulative amplitude of responses to the first 60 pulses and linearly back extrapolating through the last 20 pulses to y=0 (Fig. 4E). This also did not show a significant difference between manipulated and control animals (Fig. 4F). Though we noticed that during the 60Hz pulse train, NMJs in embryonic heat stress-manipulated specimens exhibited an increased facilitation during the first 10 pulses, indicating a slight reduction in release probability of the presynaptic terminal (Fig. 4D). Last, we tested if the NMJ remained able to adjust homeostatically to acute changes in synaptic transmission. We utilized the well-established method of acutely blocking glutamate receptors with Philanthotoxin (PhTx), a GluRIIA blocker, to induce presynaptic homeostatic potentiation (Frank et al., 2006; Genç and Davis, 2019; Müller and Davis, 2012). Acute GluRIIA blockade by a 10 min incubation with PhTx caused a significant reduction mEPSC amplitude while evoked EPSC amplitude remained unchanged in both controls and heat stress-manipulated specimens. Calculating quantal content shows that in response to acute GluRIIA blockade, both control and manipulated animals are indeed able to induce presynaptic homeostatic potentiation by increasing quantal content (Fig. 4G-Giii). While establishing baseline conditions, we found mEPSC amplitudes and quantal content to be comparable between controls and experimentals Fig. 4G-Giii). This was contrary to expectation considering the overall reduction in the GluRIIA/B ratio that results from an embryonic heat stress manipulation. Overall, we conclude that despite the significant structural changes that emerge by the late larval stage following an embryonic 32°C heat stress experience, synaptic physiology at the late larval NMJ remains normal. Moreover, when considering the unstable network phenotype, as manifest by increased recovery times from electroshock-induced seizures, our finding that synaptic transmission at the NMJ remained normal suggests that the network instability likely originates from changes within the central locomotor circuitry.

Physiological properties of the late larval NMJ following an embryonic 32°C perturbation.

A) Representative traces of control two electrode voltage clamp recordings across a gradient of external calcium concentrations. B) Postsynaptic current amplitude is not altered between control and embryonic heat stress-manipulated animals at any given external calcium concentration (n=5). C) Plotting postsynaptic current amplitude against its variance to calculate quantal size (q) and the number of release sites being used (N), also did not show a significant difference between control and embryonic heat stress-manipulated animals (n=5). D) To measure readily releasable pool size, a 60Hz stimulus train was applied for 1s where we could observe a slight increase in facilitation over the first 10 pulses in specimens that in their embryonic stage had experienced heat stress of 32°C (n=9) as compared to controls (n=11). E) Cumulative amplitude was plotted over time and a linear regression was back extrapolated through the last 20 pulses (same data as for E). F) The y intercept was divided by the Mini size of the same trace to calculate the readily releasable pool (RRP) size. G) NMJs in heat stress-manipulated specimens have normal EPSC and mEPSC amplitudes and quantal content. Acute application of the GluRIIA blocker PhTx reliably induced presynaptic homeostatic potentiation in controls and 32°C manipulated animals. Data in F & G are shown with mean ± SEM, ANOVA, *p < 0.01, **p < 0.001, ***p < 0.0001, ‘ns’ indicates statistical non-significance.

Motoneurons display reduced excitability and receive increased synaptic input following embryonic perturbation by heat stress

We noted that following embryonic exposure to 32°C heat stress, motor axons in late third instar larvae required an increased stimulation voltage to reliably induce action potentials. This hinted at motoneurons in these larvae having acquired a state of reduced excitability. To directly assay motoneuron excitability, we used whole cell patch clamp to measure action potential firing of aCC, evoked by injection of current steps (4pA). These measures were made in the presence of mecamylamine (200 µM in external saline) to block endogenous cholinergic excitatory synaptic drive. We found that in late third instar larvae, the aCC motoneuron had significantly decreased excitability following an embryonic heat stress experience of 32°C (Fig. 5A). These results are comparable to, and validate, results obtained after optogenetic locomotor network activity manipulation during the embryonic CP of the locomotor network (17-19 hrs AEL) (Giachello and Baines, 2015). In previous studies, we had shown with recordings from aCC that its synaptic drive is enhanced following transient network activity manipulations during the embryonic CP, using optogenetic or pharmacological means to induce network over-excitation or genetic persistent overactivation of neurons, as in parabss seizure mutants (Giachello and Baines, 2015; Hunter et al., 2024). This manifests as an increase in the duration (though not amplitude) of endogenous excitatory cholinergic synaptic currents, termed spontaneous rhythmic currents (SRCs) (Giachello et al., 2019; Giachello and Baines, 2015). We asked if heat stress experienced during embryonic development led to similar changes. Indeed, in late larvae that had experienced heat stress during embryonic development, SRCs recorded from aCC motoneurons were significantly increased in duration, though with amplitude unchanged (Fig. 5Bi), i.e. phenocopying the lasting effects of network activity manipulations during the embryonic CP. Note these SRC recordings were undertaken in the absence of mecamylamine and suggest an increase in premotor network activity.

Larval motoneurons display reduced excitability and increased synaptic input after an embryonic 32°C heat stress experience.

A) Embryos were subjected to 32°C heat stress, then reared to the late larval stage under control conditions of 25°C. Whole cell patch clamp recordings show action potentials generated by aCC motoneurons isolated from excitatory synaptic input by mecamylamine addition in response to current injection. Inset shows linear regression centred around 32-72 pA current injections, showing significant reduction in excitability in heat stress manipulated specimens relative to controls. B) Whole cell patch clamp recordings from the aCC motoneuron in larvae: Bi) reveal an increase in the duration of spontaneous rhythmic currents (SRCs) in larvae that had previously experienced 32°C heat stress during embryonic development. Current density and frequency of SRCs remain comparable to controls. Bii) Targeted optogenetic activation (using Chronos, for 1 second) of one premotor excitatory interneuron, A27h, showed A27h synaptic input has increased as a result of a previous embryonic 32°C heat stress experience. Data are shown with mean ± SEM, unpaired t-test, *p < 0.01, ****p < 0.00001, ‘ns’ indicates statistical non-significance.

To further explore changes in the premotor network, we recorded synaptic drive to aCC under whole cell patch clamp configuration. We focused on one of the most the strongly connected excitatory premotor interneurons, termed “A27h”. Because motoneurons are quiescent without synaptic drive, a small amount of current was injected into aCC to maintain a modest and consistent mean firing frequency of ∼10-20 Hz. We then assayed how this mean firing frequency changed upon optogenetic activation of the A27h premotor interneuron (Hunter et al., 2024). As might be expected, and as per our previous study (Hunter et al., 2024), acute optogenetic activation of the A27h premotor excitor resulted in increased aCC firing frequency. Significantly, this increase was greater in larvae that had been exposed to heat stress during their embryonic development, as compared to controls (Fig 5.Bii). This observation demonstrates that heat stress during embryonic development results in increased excitatory drive from central premotor interneurons, and it suggests that this might then countered by the postsynaptic motoneurons through reduction in their excitability.

Adjustments of the pre-motor synaptic drive onto the aCC motoneuron result in normal endogenous firing patterns

If the reduction in excitability displayed by the aCC motoneuron was a homeostatic response to the increased premotor synaptic drive, then one might predict this to be a mechanism to maintain the endogenous level of motoneuron activity. To test this, we used cell-attached patch recordings from aCC motoneurons, in isolated nerve cords of late third instar larvae.

This extracellular recording mode allows quantification of endogenous action potential firing without disturbance of the intracellular composition of a neuron. As hypothesized, our recordings show that no significant differences in endogenous motoneuron activity result from an embryonic experience of heat stress, as measured by burst rate, burst duration, or action potentials per burst (Fig. 6A). Next, to test if this translates to normal muscle output, we recorded spontaneous synaptic drive from muscles in a semi-intact preparation with an intact nervous system attached. Current clamp recordings from the DA1 muscle, while supressing muscle contractions using nifedipine (50µM) (Dyson et al., 2022; Kratschmer et al., 2021), showed no significant differences in spontaneous burst frequency or duration, nor the overall time spent bursting (Fig. 6B), between controls and embryonic heat stress-manipulated specimens. Therefore, we conclude that, at least in this locomotor system, CP manipulations might act primarily on the cells of the premotor circuitry, namely excitatory premotor interneurons as tested in this study. Changes in the motoneurons could thus be seen to constitute secondary homeostatic adjustments in response to those premotor network changes.

Spontaneous larval motoneuron and muscle firing properties are unchanged following embryonic heat stress.

A) Monitoring spontaneous motoneuron firing by cell-attached patch clamp recordings from the aCC motoneuron in a late third instar larval nerve cord found no significant differences between embryonically manipulated specimens and controls, regarding burst rate per second, burst duration or number of spikes per burst (n=10). B) Measuring muscle output generated by motoneuron spontaneous firing with current clamp recordings from the DA1 muscle in late third instar larvae showed that larvae are able to produce normal motoneuron firing output to the muscle, irrespective of embryonic experience (n=8 for controls, n=6 for experimentals with 32°C embryonic experience). Data are shown with mean ± SEM, unpaired t-test, ‘ns’ indicates statistical non-significance.

Propagation speed of neuronal activation waves is reduced following embryonic heat stress perturbation, mirroring reduced crawling speed

To establish how the excitatory premotor network is affected by a transient embryonic 32°C heat stress manipulation we conducted functional imaging of the segmentally repeated A27h premotor interneurons. Specifically, we targeted expression of the calcium indicator GCaMP8f to A27h premotor interneurons. Then, in isolated third instar nerve cords, we recorded the propagation of neuronal activation waves, indicative of fictive locomotion, across abdominal segments (Fig. 7A) (Streit et al., 2016; Pulver et al., 2015). These experiments show that embryonic 32°C exposure leads to a significant reduction in the speed with which these neuronal activity waves travel from posterior to anterior abdominal segments, as compared to controls, and mirrors the decrease in larval crawling speed caused by the same embryonic manipulation (Fig. 7B). Because our electrophysiological analyses showed that motoneuron firing patterns and output to the muscle remain normal after an embryonic 32°C manipulation (Fig. 4-6), these results further support the hypothesis that it is the lasting changes to the central premotor circuitry, rather than to the motoneurons, which are responsible for the behavioural changes, in crawling behaviour, and likely also for network instability.

Embryonic heat stress manipulation causes reduced speed of both, activity wave propagation and larval crawling.

A) Functional imaging of A27h premotor interneurons using GCaMP8f in isolated CNSs from late larvae that previously had experienced transient 32°C heat stress during embryonic development. Activity wave propagation across abdominal segments is slowed down, i.e. intersegment duration has increased, notably between anterior segments, A3 to A1. B) In intact animals, locomotor network output is similarly altered: embryonic heat stress (32°C) leads to a significant reduction in larval crawling speed, as compared to controls (25°C). Data are shown with mean ± SEM, unpaired t-test, **p < 0.001, ****p < 0.00001, ‘ns’ indicates statistical non-significance. For B) only the crawling speed of each larva was sampled up to three times, once per unique 5-minute crawling interval.

The CP of the locomotor network determines behavioural outcomes

Having identified consecutive phases of embryonic development that represent CPs for the body wall muscles and NMJ development, followed by a CP for the central locomotor circuitry (Fig. 3), we asked which of these CPs shaped larval crawling behaviour. To this end, we subjected embryos to 32°C heat stress at different windows of their development, separating the periods that principally affects NMJ development (prior to 17 hours AEL) from the subsequent 2-hour window during which the development of the locomotor circuitry is vulnerable to neuronal activity and heat stress manipulations (17-19 hours AEL) (Giachello and Baines, 2015). Assaying crawling behaviour of third instar larvae showed that crawling speed is reduced only when the embryonic CP of the locomotor network (17-19 hours AEL) had been subjected to heat stress. This demonstrates that it is the impact on the developing central circuitry, rather than NMJ formation, which causes behavioural changes in terms of crawling speed (Fig. 8A) and of network stability, as assessed by seizure recovery times (Fig. 1).

Larval crawling speed is determined by the CP of central locomotor network.

A) Embryos were subjected to heat stress (32°C) during different windows of development. Crawling speed of resultant third instar larvae was affected only when the CP for the central locomotor network (17-19hrs after egg laying [AEL]) was manipulated. B) Embryonic experience of heat stress leads to larvae that retain their ability to increase crawling speed in response to acute relative increases in ambient temperature, to a higher absolute temperature (35°C) than controls (29°C). Data are shown with mean ± SEM, ANOVA, *p < 0.01, **p < 0.001, ****p < 0.00001, ‘ns’ indicates statistical non-significance. Each larva was sampled up to three times, once per unique 5-minute crawling interval.

We wondered whether larvae that emerge from heat stressed embryos crawl more slowly simply because they are no longer able to crawl at the normal speed. To test this, we subjected embryonic development to 18, 25 or 32°C and, after hatching, reared larvae at 25°C until the mid-third instar stage (72 hrs ALH). We then assayed how these larvae adjusted their crawling speed in response to an acute change in ambient temperature, when placed into an arena equilibrated to 18, 25, 29, 32 or 35°C (Fig. 8B). This clearly showed that while embryonic heat stress leads to a significant reduction in the default crawling speed, larvae retained the ability to accelerate in response to an acute increase in ambient temperature, as they normally do. Moreover, animals that experienced a higher temperature during embryogenesis retained the ability to increase their crawling speed at higher temperatures (e.g. 32 or 35°C) than those with cooler embryonic experiences, whose maximum speed plateaued at 29°C (Fig. 8B). These data suggest that while an embryonic experience of heat stress reduces the default crawling speed, it preserves a response range at higher temperatures, which might therefore potentially be adaptive to larval life at higher temperatures.

Discussion

An experimental model system with which to study CP biology

In this study we utilised the Drosophila embryo and larva as an experimental model with which to study the role of CP-regulated plasticity across interconnected cells within a defined circuit, from a premotor excitatory interneuron, to motoneuron to muscle. We have identified 32°C heat stress as a CP stimulus that has ecological relevance, as a temperature experienced in the wild (Williamson et al., 2021). Transient exposure of embryos to 32°C during the previously defined CP for the developing CNS (17-19 hours after egg laying) leads to lasting changes in subsequent nervous system development, resulting in decreased crawling speed of third instar larvae and in reduced network stability, reflected by an increased susceptibility to electroshock-induced seizures. As is characteristic for CPs, the same manipulation before or after this defined developmental window has no lasting effect on nervous system properties or animal behaviour. A well-known way by which increases in temperature impact on neurons is their activity. Using calcium imaging of motoneurons we showed that stepwise increases in ambient temperature cause acute increases of fictive rhythmic locomotor network activity in isolated nerve cords. This agrees with past studies, which have demonstrated the well-established link between temperature and neuronal activity in this same experimental system (Zhong and Wu, 2004). It is therefore not surprising that 32°C heat stress during the embryonic CP of nervous system development causes outcomes that are comparable to those created by neuronal activity manipulations (pharmacological or genetic) during this developmental window (Giachello et al., 2021; Giachello and Baines, 2015; Hunter et al., 2024).

Different cell types respond differently to CP manipulations

A particular strength of the Drosophila larval motor system is the accessibility of interconnected network elements. We focused on a small part of the locomotor network: an excitatory premotor interneuron (A27h) connected to a motoneuron (aCC), and its postsynaptic body wall muscle (DA1). Following transient application of heat stress during embryonic development, we could identify that within this section of the locomotor network each of these three connected nodes responded differently. We observed increased excitatory drive from A27h interneurons to aCC motoneurons, which might reflect changes in A27h neuronal excitability and/or its synaptic transmission onto motoneurons. Potentially in response to these changes in the premotor network, aCC motoneurons displayed a decrease in excitability, which facilitated normal motoneuron firing patterns to be maintained. In contrast to these changes within the central circuitry, at the peripheral NMJ terminal, we documented normal synaptic transmission properties, despite a significant presynaptic terminal overgrowth. The latter might be a secondary adjustment to a decrease in transmitter sensitivity by the muscle, which showed reduced levels of the high conductance subunit GluRIIA.

The first observation to make is that these responses to embryonic heat stress, which manifest during subsequent larval stages, are reproducible. This suggests that there might be direct cause-effect relationships at the cellular level, which can now be investigated in detail using this experimental model. Building on these findings, we thus identified mitochondrial reactive oxygen species (ROS) and Hypoxia-inducible factor-1 alpha (HIF-1α) as heat stress-activated signals required for the long-lasting changes of cellular properties that result from embryonic CP heat stress (Sobrido-Cameán et al., 2025). An alternative, not mutually exclusive interpretation is that activity patterns in this part of the network might be robustly canalised so as to lead to reproducible outcomes. Second, the changes in neuronal excitability and synaptic transmission that we have documented are compatible with a temporal sequence, or a functional hierarchy, during network maturation. For example, following an embryonic heat stress, premotor interneurons, such as A27h, potentially part of the central pattern generator (Carreira-Rosario et al., 2021; Zeng et al., 2021), increase their collective synaptic drive to postsynaptic motoneurons, which in turn reduce their excitability (Fig. 5). Along these lines we interpret the changes in motoneuron properties that we have characterised as homeostatic responses by motoneurons to the increase in presynaptic excitatory drive from premotor interneurons. This perspective implies a hierarchy, whereby the specification of premotor interneuron properties dominates that of their downstream integrators, the motoneurons (Fig. 9). This is supported by previous findings of motoneurons flexibly adjusting their electrical properties in response to changes in presynaptic excitatory drive (Baines et al., 2001). However, the precise temporal and causal relationships of cell type-specific adjustments during the CP of this network remain to be established. In the developing mouse neocortex, sequential cell type-specific maturation windows have recently been shown. There, somatostatin-expressing inhibitory interneurons differentiate early, and promote the maturation of network activity patterns through inhibition of the later differentiating parvalbumin-expressing inhibitory interneurons (Mòdol et al., 2024).

Model of distinct critical periods for body wall muscles and the central locomotor circuitry.

Embryonic heat stress (32°C) leads to adjustments of postsynaptic receptor fields, when applied during an earlier developmental window of 13-16 hrs after egg laying. Subsequently, from 17-19 hrs after egg laying the central locomotor circuitry is sensitive to perturbations, leading to an increase in synaptic drive of the premotor network, potentially including central pattern generating circuits (CPGs). In response, motoneurons decrease their excitable properties, thus maintaining normal locomotor output to the body wall muscles. Nevertheless, changes in the premotor network lead to a reduction in activity wave progression and thus reduced larval crawling speed, also manifesting a reduction in overall network stability.

NMJ structural development is affected by CP manipulations, but not synaptic transmission

Despite embryonic heat stress leading to significant NMJ structural changes, in terms of presynaptic terminal overgrowth (an average increase of active zone number by 30%) and overall reduction of GluRIIA (by 50%), evoked transmission resulting from the firing of single action potentials in the motor axon remained normal. Distal boutons had previously been shown to house release sites that are preferentially recruited under low frequency stimulation regimes, such as used for making NMJ recordings (Peled and Isacoff, 2011). Speculatively, preferential maintenance of GluRIIA in distal boutons, might have explained maintenance of normal synaptic transmission at low frequency stimulation regimes. However, we found no evidence for this. On the contrary, in embryonic heat stress-manipulated specimens, GluRIIA levels were comparably reduced in both proximal and distal boutons (data not shown). An alternative explanation for maintained synaptic transmission in the face of increased active zone number is a reduction in release probability. Although we did not conduct functional imaging to investigate this directly, an indirect indication for this is the increased facilitation in animals that had experienced embryonic heat stress, which indicates a reduction in presynaptic release probability across the whole NMJ (Fig. 4D).

On reflection, to find evoked synaptic transmission at the NMJ not significantly affected by CP manipulations, is perhaps not unexpected. A well-established feature of this particular synapse is that its transmission is tightly regulated so as to maintain constancy through the deployment of multiple NMJ plasticity mechanisms, including presynaptic homeostatic potentiation (PHP), which we have shown to function effectively even after an embryonic heat stress manipulation (Fig. 4G) (Delvendahl and Müller, 2019; Frank et al., 2020, 2006; Goel and Dickman, 2021).

Distinct developmental windows with sensitivity to heat stress suggest sequential network maturation

It is unlikely that structural changes at the NMJ contribute to making the nervous system less stable, or that they effect CP-regulated changes to larval behaviour, such as crawling speed or social feeding (Williamson et al., 2021). Based on previous CP manipulations targeted to various neuronal subsets (Giachello et al., 2021), it seemed more likely that behavioural phenotypes, including reduced crawling speed and increased recovery times from electroshock-induced seizures, would be the consequence of changes in the central circuitry. Indeed, we found that the developing body wall muscles have an earlier CP that is sensitive to heat stress (13-16 hrs AEL) and leads to changes in postsynaptic GluR composition (Fig. 3), than the central locomotor circuitry (17-19 hrs AEL) (Figs, 1 & 8A). By applying heat stress manipulations to these distinct developmental phases we were able to demonstrate that behavioural phenotypes on both larval crawling speed and network stability are linked to CP-induced changes to the central circuitry, not the peripheral NMJ (Figs. 1, 3 & 8A). These CPs correlate with the developmental phases when the body wall muscles and central neurons become electrically active, respectively (Broadie and Bate, 1993; Baines and Bate, 1998). Thus, they reflect the sequence of earlier myogenic and subsequent neuronal maturation (Crisp et al., 2008). This is comparable with observations from mammalian systems, where different parts of the developing brain mature in sequence (Hensch, 2005; Hensch and Quinlan, 2018; Reh et al., 2020).

Mechanistically, heat stress causes increased neuronal activity (Oswald et al., 2018; Sigrist et al., 2003; Zhong and Wu, 2004). Functional imaging of motoneurons showed that in larvae acute temperature elevations cause corresponding increases in both fictive locomotor activity waves (Fig. 1B) as well as larval crawling (Fig. 8B). Interestingly, this manipulation, when carried out during the embryonic CP, leads to a locomotor network with significantly slower intersegmental propagation of activity waves and, similarly, reduced crawling speed (Fig. 7). One possibility is that heat stress during the CP might cause a lasting malfunctioning of pacemaker activity. Alternatively, the slower central pattern of locomotor network activity in larval stages could be the consequence of a network-wide homeostatic response to an abnormally fast rate provoked by transient heat stress during the embryonic CP. The latter explanation is compatible with our recent findings of the CP as an integration window of excitatory vs inhibitory network activity (Hunter et al., 2024) and the identification of nitric oxide signalling pathways downstream of CP network activity (Giachello et al., 2021). To identify additional genetic pathways associated with CP plasticity, we propose that the NMJ, as a large, readily accessible and morphometrically quantifiable peripheral synapse, could serve as a convenient and effective assay.

Materials and Methods

Fly husbandry

Wild-type and transgenic strains were maintained on standard yeast–agar–cornmeal medium at 25°C. For CP manipulations, animals were kept in laying pots with apple juice agar plates and yeast paste for 6h. Collected eggs were either kept at 25°C for controls or at 32°C for 24h, before being moved back to 25°C until animals developed to the L3 larval stage.

Dissection and Immunostainings

Wandering L3 larvae were dissected in Sørensen’s sodium phosphate buffered saline (0.1 M, pH 7.2) and fixed for 10min at room temperature with Bouińs fixative (Sigma-Aldrich). After fixing, specimens were washed in washing solution (Sørensen’s buffer + 0.25% BSA + 0.3% TritonX-100 from Sigma-Aldrich) and incubated with mouse nc82 (anti-Bruchpilot to reveal active zones) antibody (1:100, DSHB) overnight, visualized with goat α-mouse StarRed (1:1000, Abberior, cat.no.STRED-1001-500UG, Lot. 2042PK-1) and AlexaFluor488 conjugated goat α-HRP (1:1000, Jackson ImmunoResearch, cat.no. 123-545-021, Lot. 152767) to visualise presynaptic, neuronal structures. For stainings of the postsynaptic glutamate receptor subunits GluRIIA and GluRIIB we used monoclonal mouse-α-DGluRIIA antibody (1:100, DSHB, product ID: 8B4D2 (MH2B)) and polyclonal rabbit-α-DGluRIIB antiserum (1:1000, gift from Mihaela Serpe (Sulkowski et al., 2016)) in conjunction with goat α-mouse StarRed (1:1000) and goat α-rabbit Atto594 secondary antibodies (1:1000, Hypermol, cat.no. 2306, Lot. 0356RG22), respectively, in combinations with AlexaFluor488 conjugated goat α-HRP (1:1000, Jackson ImmunoResearch, cat.no. 123-545-021, Lot. 152767) to visualise presynaptic terminals.

Fly stocks and other Key Resources

Image acquisition

Confocal images were acquired, using a Leica Stellaris point scanning confocal microscope, with an 40x/1.25N.A. oil immersion lens (Leica) and a 1.25 digital zoom. Images were acquired with a voxel size 0.16x 0.16x 0.3 µm. Images were processed using FIJI and Affinity Designer.

Two electrode voltage clamp

Two Electrode Voltage Clamp (TEVC) recordings were performed in HL3.1 saline (Feng et al., 2004) composition in mM: NaCl 70, KCl 2.5, MGCl2 4, CaCl2 2 (if not other mentioned), NaHCO3 10, trehalose 5, sucrose 115, HEPES 5 (pH was adjusted to 7.24-7.25 with 1M NaOH), using an Olympus BX50WI compound microscope with 60X dipping (Olympus 60x/0.9 N.A.) objective lenses. Experiments were performed on third-instar larval NMJs of the dorsal acute muscle 1 (DA1) in segments A3 and A4. The respective nerve was sucked in with a self-built suction electrode (Johnson et al., 2007) which was filled with HL3.1 saline. Subsequently, the nerve was stimulated for 0.1ms to trigger action potential firing. Briefly, voltage was stepwise increased until the first synaptic transmission could be recorded. From that voltage, 1 V was added to reliably evoke APs in control conditions, in 32°C CP manipulated animals 2-3 V had to be added to reliably recruit aCC and RP2. For PhTx experiments, samples were incubated in 4 µM PhTx final concentration (Frank et al., 2006) for 10 min. While dissecting, great care was taken to not stretch animals too much. Successful PhTx application was determined by dividing miniature EPSC amplitude of PhTx treated animals by the average Amplitude of control mEPSCs. If PhTx treated mEPSCs were at least 40% reduced compared to control, PhTx application was considered successful (Frank et al., 2006). Quantal content was determined by dividing EPSC by mEPSC for each recording. Recordings were only further considered from cells with an initial Vm below -50 mV and input resistance between 8-10 MΩ, using intracellular electrodes filled with 3M KCl with resistances of ∼20 MΩ (Electrodes for current injection and the membrane voltage). Current passing und recording electrodes were pulled from Borosilicate glass microelectrodes with filament (OD1 mm, ID 0.58mm; Harvard Apparatus LTD, GC100F-10 Part No. 30-0019), suction electrodes from Borosilicate glass microelectrodes without filament (OD1 mm, ID 0.58mm; Harvard Apparatus LTD, GC100-10 Part No. 30-0016) with a Flaming/Brown Puller Model P-97 (Sutter Instruments USA). Excitatory postsynaptic currents (EPSCs) were evoked by an Isolated Pulse Stimulator (Model 2100 A-M Systems USA), recorded at a clamped voltage of -60 mV with a sampling rate of 50 kHz and filtered with the anti-alias filter at 25 kHz using the AxoClamp 2B amplifier (Molecular Devices Axon Instruments USA). Data analysis was done with Clampfit 10.7. Traces were first filtered with a 360 Hz Gaussian low pass filter, then template searches were run over the trace for single pulses and spontaneous miniature excitatory postsynaptic currents (mEPSCs or minis). The resulting files were used to analyse mEPSC and EPSC properties. Rise and decay tau were measured with an one-term exponential product fit (Levenberg-Marquadt method with a max of 5000 iterations)

Current clamp muscle recordings

Current clamp recordings from muscle DA1 were performed in HL3.1 saline (see above), in semi intact, fillet dissected wandering L3 larvae with intact nervous system. After finishing the dissection, only animals performing crawling-like movements were considered for further experiments. To block muscle contractions, a 500 µM stock solution of nifedipine in HL3.1 and 1% DMSO was added to the bath for a final concentration of 50 µM nifedipine and 0.1% DMSO for 15 min. Recordings were only further considered from cells with an initial Vm below -60 mV and input resistance between 8-10 MΩ, using intracellular electrodes filled with 3M KCl with resistances of ∼20 MΩ. Spontaneous activity was recorded for 5min. Burst analysis was performed using the Clampfit Burst Analysis, using the Poisson Surprise model with a minimum of 5 events per burst.

Electroshock assay

A handheld stimulator consisting of two tungsten wires (0.1 mm diameter) fixed to a plastic rod was used. Wires were shaped to be ∼1-2 mm apart, attached to a DS2A Mk.II stimulator (Digitimer Ltd), which was set to provide a 6 V, 2 s current pulse. Larvae were placed on a plastic plate lid, with tissue paper used to dry residual moisture from the larvae. Once larvae had returned to normal crawling, the tungsten wires were brought into contact with the larvae, positioned on the larval cuticle, perpendicular to and over the approximate position of the CNS, with gentle pressure. Current stimulation was then delivered, which usually induced a sustained contraction and paralysis representing a seizure (videos can be viewed in Huertas Radi et al., 2025 and Marley and Baines 2011). The time to resumption of normal behaviour (regular, whole body forward propagating peristaltic waves) was recorded and defined as the duration of recovery from seizure.

Dissection and CNS mounting for calcium imaging and electrophysiology

Larval CNSs were dissected and mounted for electrophysiological recordings and for GCaMP imaging. Third instar (L3) larvae (of either sex) were dissected in a dish in standard saline (135 mM NaCl (Fisher Scientific), 5 mM KCl (Fisher Scientific), 4 mM MgCl2·6H2O (Sigma-Aldrich), 2mM CaCl2·2H2O (Fisher Scientific), 5 mM TES (Sigma-Aldrich), 36 mM sucrose (Fisher Scientific), pH 7.15), to remove the CNS (ventral nerve cord and brain lobes). The isolated CNS was then transferred to a droplet of external saline, within which it was laid flat (dorsal side up) and glued (GLUture Topical Tissue Adhesive; World Precision Instruments USA) to a Sylgard-coated cover slip (1 to 2mm depth of cured SYLGARD Elastomer (Dow-Corning USA) on a 22 x 22mm square coverslip). The preparation was then placed on a glass slide and viewed under a microscope (Olympus BX51-WI).

Calcium imaging

Dissected CNSs were imaged (QImaging EXi-Aqua; photometrics, Arizona, US) with an acquisition frame rate of 10Hz and frame duration of 100 ms using a x20 immersion lens. GCaMP8f was excited using a 470 nm collimated LED (Thorlabs, New Jersey, US). Time-course of neuronal fluorescence was determined in WinFluor (V.4.1.5; University of Strathclyde, UK) by adding regions of interest (ROIs) to axons, close to the cell body (Streit et al, 2016). In Clampfit (10.3.1.5; Molecular Devices, California, US) fluorescence traces were smoothed using a lowpass BoxCar filter using 11 smoothing points prior to analysis.

For motoneuron calcium transients (Fig. 1), recordings were conducted for 600s, whilst temperature of the explanted CNS was increased steadily from ∼23°C to ∼34°C using a digital heating unit (CO 102; Linkam Scientific, Surrey, UK). For sham recordings, temperature was maintained at a constant 25°C. Temperature was measured in Clampex (10.3.1.5; Molecular Devices, California, US) using a temperature sensor placed in the external saline surrounding the brain, via a temperature controller (TC-10; npi electronic GmbH, Tamm, Germany). The frequency of calcium fluorescence peaks was measured for each segment (A8/9 – A1) in three sequential 200 s time periods and averaged per larva. Average temperature was calculated for each recording during these same 200 s time periods.

For calcium imaging of interneurons after CP manipulation (Fig. 5), 180 s recordings were taken. Time lag was determined by manually assessing the time between peak fluorescence corresponding to ipsilaterally adjacent interneurons (segments A6 – A1. For each recording the first complete forward waves of activity were used (up to three repeats per trace) to generate an average intersegment duration for each pair of interneurons.

Whole cell patch clamp electrophysiology

CNSs were dissected, mounted (see calcium imaging methods) and viewed under a 60x water-immersion lens. To access soma, 1% protease (Streptomyces griseus, Type XIV, Sigma-Aldrich, in external saline) contained within a wide-bore glass pipette (GC100TF-10; Harvard Apparatus UK, approx. 10 µm opening) was applied to abdominal segments, roughly between A5-A2 (Baines and Bate, 1998). This was done to remove overlaying glia to facilitate access to underlying nerve cell soma (e.g., aCC). Motoneurons were identified by anatomical position and relative cell size, with aCC being positioned close to the midline and containing both an ipsilateral and contralateral projection. Whole-cell patch recordings were made using borosilicate glass pipettes (GC100F-10, Harvard Apparatus) that were fire polished to resistances of 10 - 15MΩ when filled with intracellular saline (140 mM potassium-D-gluconate (Sigma-Aldrich), 2 mM MgCl2·6H2O (Sigma-Aldrich), 2 mM EGTA (Sigma-Aldrich), 5 mM KCl (Fisher Scientific), and 20 mM HEPES (Sigma-Aldrich), (pH 7.4). Input resistance was measured in the ‘whole cell’ configuration, and only cells that had an input resistance ≥ 0.5 GΩ were used for experiments. Cell capacitance and break-in resting membrane potential was also measured for each cell recorded. Data for current steps recordings was captured using a Multiclamp 700B amplifier controlled by pCLAMP (version 10.7.0.3), via an analogue-to-digital converter (Digidata 1440A, Molecular Devices). Trace/s were sampled at 20 kHz and filtered online at 10 kHz.

Current step recordings

Mecamylamine (200 µM in saline) was used to block postsynaptic nACh receptors to synaptically isolate neurons for experiments designed to measure intrinsic motoneuron excitability. Once patched, neurons were brought to a membrane potential of -60 mV using current injection. Each recording consisted of 20 x 4 pA (500 ms) current steps, including an initial negative step, giving a range of -4 to +72 pA. Number of spikes were counted and plotted against injected current. Both cell capacitance and input resistance were compared between conditions to ensure that any observed differences in excitability were not due to differences in either cell size or resistance.

Synaptic drive to motoneurons

To measure synaptic drive between A27h premotor interneurons and aCC, A27h neurons were identified via Chronos-mVenus reporter (as above), allowing corresponding aCC motoneurons to be patched. Recordings were conducted in current-clamp, with current injected into aCC sufficient to evoke a mean spike frequency of ∼10-20 Hz, prior to optogenetic stimulation. Recordings consisted of: individual sweeps, beginning with 1 s LED off, followed by 1 s LED on, and 1 s LED off (repeated 5-8 times per cell). Change in AP frequency (%) was calculated for each recording (per sweep) comparing AP frequency before and during stimulation, which was then averaged for each cell.

Spontaneous rhythmic currents (SRCs) were recorded for 180 s with aCC voltage clamped at -60 mV. Amplitude, duration, and frequency were measured for the first 15 events of each trace using Clampfit’s event detection, threshold search, function. The first 15 events with a stable baseline, an amplitude of >300 pA and no double peaks were accepted for analysis. Amplitude was measured as the change from baseline immediately prior to event initiation to the peak current and normalised for cell capacitance, providing a measure of current density. SRC duration was measured as the width of an event at 100 pA amplitude.

Spontaneous motoneuron recordings

Cell-attached recordings, measuring motoneuron activity, were conducted using unpolished patch electrodes. Motoneuron spontaneous activity was recorded using an adapted cell-attached current clamp protocol (Multiclamp 700B amplifier with I=0) to measure passive action potential bursting properties. In Clampfit (version 10.3.1.5), 180 s recordings, were analysed to detect action potential ‘spike times’. These were imported into MATLAB and analysed using a custom script (developed by Mituzaite et al., 2021). An action potential burst was considered as an event consisting of at least 3 spikes occurring within 100 ms and ending when no spike was detected for 100 ms.

Crawling Assay

Crawling behaviour was recorded at the mid-third instar stage, 72 hours after larval hatching (ALH). Larvae were rinsed in water and placed inside a 24cm x 24cm crawling arena with a base of 5mm thick 0.8% agarose gel (Bacto Agar), situated inside an incubator. Temperature was maintained at 25±0.5°C, reported via a temperature probe in the agar medium. Humidity was kept constant. Larval crawling was recorded using a frustrated total internal reflection-based imaging method (FIM) in conjunction with the tracking software FIMTrack (Risse et al., 2017, 2013) using a Basler acA2040-180km CMOS camera, fitted with a 16mm KOWA-IJM3sHC.SW-VIS-NIR lens, acquisition controlled by Pylon software (Basler) and Streampix (v.6) software (NorPix). Larvae were recorded for 20 minutes at five frames per second.

Recordings were split into four 5-minute sections with the first five minutes of acclimatisation discarded. The remaining three 5-minute sections were used to analyse crawling speed, choosing crawling periods of uninterrupted forward crawling, devoid of pauses, turning or collision events. Each larva was sampled up to once per 5-minute section, i.e. up to three data points were derived per larva from up to three consecutive 5-minute segments. Crawling speed was calculated using FIMTrack software.

Statistics

Statistical analyses were done using GraphPad Prism Software (Version 10.1.2). Datasets were tested for normal distribution with the Shapiro-Wilk Test. Normally distributed data were then tested with students t-test (for pairwise comparison). Normally distributed analysis for more than two groups was done with a one-way ANOVA and post hoc tested with a Tukey multiple comparison test. Non-normally distributed data sets of two groups were tested with Mann-Whitney U Test (pairwise comparison) and datasets with more than two groups were tested with a Kruskal Wallis ANOVA and post hoc tested with a Dunns multiple post hoc comparison test. For whole cell electrophysiology experiments in isolated neurons, linear regression analysis was used to compare the intercepts between conditions. This analysis was restricted to linear sections of data, represented by inset graphs. For all datasets mean and standard error of mean (SEM) are shown. Significance levels were * p<0.05; ** p<0.01; *** p<0.001; **** p<0.0001. Sample sizes are stated in figures or legends, each data point derived from a unique specimen, unless otherwise stated; e.g. Figure 3B, NMJ sizes were sampled at up to four data points per larva; for larval crawling speeds up to three data points were derived per larva, one for each unique 5-minute segment of recorded crawling. All experiments indiscriminately included larvae of both sexes.